### 2011-11-29 paper/paper.tex
Added template for the paper, and a makefile for building it.

\documentclass[journal]{IEEEtran}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}

% declare the path(s) where your graphic files are
\graphicspath{{./pdf/}}

\begin{document}

% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Dynamic Load Balancing for\\Heterogeneous Systems}

% author names and IEEE memberships
\author{Kenneth~Lee,~Kyle~Morgan}%

% The paper headers
\markboth{CS 5504: Computer Architecture}{Final Project Paper, 5 December 2011}

% make the title area
\maketitle

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
enim ad minim veniam, quis nostrud exercitation ullamco laboris
nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
in reprehenderit in voluptate velit esse cillum dolore eu fugiat
nulla pariatur. Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.
\end{abstract}

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
\IEEEPARstart{L}{orem} ipsum dolor sit amet, consectetur adipisicing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
enim ad minim veniam, quis nostrud exercitation ullamco laboris
nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
in reprehenderit in voluptate velit esse cillum dolore eu fugiat
nulla pariatur. Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{Motivation}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
enim ad minim veniam, quis nostrud exercitation ullamco laboris
nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
in reprehenderit in voluptate velit esse cillum dolore eu fugiat
nulla pariatur. Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{Architecture Overview}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
enim ad minim veniam, quis nostrud exercitation ullamco laboris
nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
in reprehenderit in voluptate velit esse cillum dolore eu fugiat
nulla pariatur. Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.

Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
enim ad minim veniam, quis nostrud exercitation ullamco laboris
nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
in reprehenderit in voluptate velit esse cillum dolore eu fugiat
nulla pariatur. Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.

\section{Conclusion}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
enim ad minim veniam, quis nostrud exercitation ullamco laboris
nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
in reprehenderit in voluptate velit esse cillum dolore eu fugiat
nulla pariatur. Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.

\begin{thebibliography}{1}

\bibitem{label1}
A.~Lincoln and G.~Washington, \emph{Some Sweet Paper Title}, 3rd~ed. Harlow, England: Addison-Wesley, 1999.

\bibitem{label2}
J.~Smith, \emph{On Having the Most Common Name}, blah blah blah derp derp derp, 2012.

\end{thebibliography}

\newpage
\appendix[Team Member Contributions]
Kenneth Lee did A, B, C.  Kyle Morgan did X, Y, Z.
\end{document}

### 2011-11-29 paper/paper.tex
Added abstract draft.

@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+Leveraging maximimum performance from specialized hardware is one of the
+major problems facing computer scientists today.  General-purpose computing
+on graphics processing units (GPGPU) is a technique that is becoming
+increasingly popular for handling computations typically performed by the
+CPU.  However, many GPGPU algorithms place the entire workload on the GPU,
+leaving the CPU idle for the duration of the computation.  This paper
+investigates the effect various load-balancing schemes across the CPU and
+GPU have on overall application performance.  Additionally, the discrete and
+fused GPU architectures, and their impact on performance, will be investigated.
 \end{abstract}

### 2011-11-29 paper/paper.tex
Added temporary introduction

@@
 % \IEEEPARstart{T}{his demo} file is ....
-\IEEEPARstart{L}{orem} ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+\IEEEPARstart{L}{everaging} the maximum performance from specialized hardware is one
+of the major challenges computer scientists face today.  The unique hardware architecture
+of the Graphics Processing Unit (GPU) gives it a place as a massively parallel hardware
+accelerator.  However, because of the difficulty in programming for this hardware,
+most of the effort is spent on trying to achieve performance from the GPU, while
+leaving the CPU unused.  By also using the CPU to process a smaller part of the
+data we can achieve higher utilization of resources and therefore increase program
+performance.
+
+GPU+CPU co-processing hardly a new idea.  Jimenez et al. \cite{jimenez} present a
+dynamic library which can be used in a system to dynamically schedule tasks for GPU
+computation across various processes. The method describes only looks at optimizing
+usage of the GPU across the whole system, instead of optimizing a single application’s
+performance. The StarPU system, on the other hand, presents a method of co-processing
+and load balancing on a per application basis \cite{augonnet}. They show that by using an
+appropriate model for the load balancing, they are able to achieve improvements in
+speed for applications which call for code acceleration multiple times.  We intend
+to extend this work to improve the speed of each kernel execution by splitting the
+work of that kernel over the capable resources, similar to the schemes used by OpenMP.
+Both of these papers address data transfer times as a major cost of GPU computing.
+Using the new AMD Fusion architecture, which fuses the CPU and GPU onto a single die,
+we may be able to eliminate or greatly reduce this cost, which will impact the overhead
+of GPU+CPU co-processing.
+
+We expect that by using GPU+CPU processing we will improve the performance of the
+application, when compared to CPU-only or GPU-only.  We will measure performance of the
+application using a wall time clock.  We also hypothesize that the fused GPU+CPU
+architecture will allow for lower overhead for load balancing.
 
@@
 
-\bibitem{label1}
-A.~Lincoln and G.~Washington, \emph{Some Sweet Paper Title}, 3rd~ed. Harlow, England: Addison-Wesley, 1999.
+\bibitem{augonnet}
+C.~Augonnet, S.~Thibault, R.~Namyst, and P.A.~Wacrenier,
+``StarPU: A Unified Platform for Task Scheduling on Heterogeneous Multicore Architectures'' in
+\emph{Proceedings of the 15th International Euro-Par Conference on Parallel Processing}, 2009, pp. 863-874.
 
-\bibitem{label2}
-J.~Smith, \emph{On Having the Most Common Name}, blah blah blah derp derp derp, 2012.
+\bibitem{che}
+S.~Che, M.~Boyer, J.~Meng, D.~Tarjan, J.W.~Sheaffer, S.H.~Lee, and K.~Skadron,
+``Rodinia: A benchmark suite for heterogeneous computing'' in
+\emph{Proceedings of the 2009 IEEE International Symposium on Workload Characterization (IISWC)}, 2009, pp. 44-54.
+
+\bibitem{danalis}
+A.~Danalis, G.~Marin, C.~McCurdy, J.S.~Meredith, P.C.~Roth, K.~Spafford, V.~Tipparaju, and J.S.~Vetter,
+``The Scalable Heterogeneous Computing (SHOC) benchmark suite'' in
+\emph{Proceedings of the 3rd Workshop on General-Purpose Computation on Graphics Processing Units}, 2010, pp. 63-74.
+
+\bibitem{jimenez}
+V.J.~Jiménez, L.~Villanova, I.~Gelado, M.~Gil, G.~Fursin, and N.~Navarro,
+``Predictive Runtime Code Scheduling for Heterogeneous Architectures,'' in
+\emph{Proceedings of the 4th International Conference on High Performance Embedded
+Architectures and Compilers}, 2009, pp. 19-33.
 

### 2011-11-29 paper/paper.tex
Added some architecture diagrams.
Added related works section to paper

--- 2011-11-29 paper/paper.tex
@@
 \section{Architecture Overview}
+In this section we discuss the different architectures of the experimental
+machines under test.  We will discuss CPU, GPU, and APU architectures in
+terms of major differences that affect the performance of the device.
+
+\subsection{CPU Architecture}
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
@@
 
+\subsection{GPU Architecture}
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
@@
 sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\subsection{APU Architecture}
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\section{Related Work}
+While there has been a lot of research done for accelerating applications
+using the GPU, most of them focus on only using the GPU for acceleration
+instead of co-processing with the CPU.
+
+Ryoo et al. discuss optimization strategies for GPU versus CPU runtime
+systems~\cite{Ryoo2007}.  However, their research is based on a GPU-only
+or CPU-only implementation instead of a heterogeneous approach.  The work
+presented in ~\cite{Jablin2011} describes a runtime and compiled framework
+for improving memory access latency and reducing data transfer overhead
+for GPU kernels.  Our work differs from this, because the kernel execution
+was still being performed soley on the GPU, whereas our work investigates
+kernel execution on multiple devices.
+
+Computation via heterogeneous systems is a relatively new research area for
+high performance computing.  The various frameworks and large number of
+different accelerator devices can make this quite difficult for application
+programmers to manage.  The StarPU system~\cite{Augonnet2009} presents a 
+framework to allow for programmers to write numerical codelets and have them
+automatically run on a heterogeneous system.  Scheduling work for this system
+is one of the most difficult parts of the system, because of the various
+overheads for work-sharing schemes. We simplify this work, by looking at a
+two-device system, and focus on the performance characteristics of different
+scheduling scheme.  Verner et al. use this approach to create a high-throughput
+low-lantecy heterogeneous encryption system~\cite{Verner2011}.
+
+Jimenez et al.~\cite{Jimenez2009} present a dynamic library which can be used
+in a system to dynamically schedule tasks for GPU computation across various
+processes. The method describes only looks at optimizing usage of the GPU across
+the whole system, instead of optimizing a single application's performance.
 

### 2011-11-29 paper/paper.tex

--- 2011-11-29 paper/paper.tex
@@
 % declare the path(s) where your graphic files are
-\graphicspath{{./pdf/}}
+\graphicspath{{./diagrams/}}
 
@@
 sunt in culpa qui officia deserunt mollit anim id est laborum.
+Blah blah blah see figure \ref{fig:cpu_arch}.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{cpu_architecture}
+\caption{CPU Architecture}
+\label{fig:cpu_arch}
+\end{figure}
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.5in]{gpu_architecture}
+\caption{GPU Architecture}
+\label{fig:gpu_arch}
+\end{figure}
+
+\subsection{APU Architecture}
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\section{Related Work}
+Ryoo et al. discuss optimization strategies for GPU versus CPU runtime
+systems~\cite{Ryoo2007}.  However, their research is based on a GPU-only
+or CPU-only implementation instead of a heterogeneous approach.  The work
+presented in ~\cite{Jablin2011} describes a runtime and compiled framework
+for improving memory access latency and reducing data transfer overhead
+for GPU kernels.  Our work differs from this, because the kernel execution
+was still being performed soley on the GPU, whereas our work investigates
+kernel execution on multiple devices.
+
+Computation via heterogeneous systems is a relatively new research area for
+high performance computing.  The various frameworks and large number of
+different accelerator devices can make this quite difficult for application
+programmers to manage.  The StarPU system~\cite{Augonnet2009} presents a 
+framework to allow for programmers to write numerical codelets and have them
+automatically run on a heterogeneous system.  Scheduling work for this system
+is one of the most difficult parts of the system, because of the various
+overheads for work-sharing schemes. We simplify this work, by looking at a
+two-device system, and focus on the performance characteristics of different
+scheduling scheme.  Verner et al. use this approach to create a high-throughput
+low-lantecy heterogeneous encryption system~\cite{Verner2011}.
+
+Jimenez et al.~\cite{Jimenez2009} present a dynamic library which can be used
+in a system to dynamically schedule tasks for GPU computation across various
+processes. The method describes only looks at optimizing usage of the GPU across
+the whole system, instead of optimizing a single application's performance.
 

### 2011-11-29 paper/paper.tex
Added basic CPU and GPU architecture subsections

@@
 \subsection{CPU Architecture}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+The CPU architecture is designed as a low latency architecture.  To facilitate
+the needs for low latency application performance, a series of hierarchical 
+caches have been designed to hide memory access costs, the most common cause
+of latency in single-threaded applications on modern architectures.  Each core
+of the CPU typically only runs one or two hardware-enabled threads to reduce
+cache pollution, thus avoiding latency.  An example CPU architecture is given
+by Figure \ref{fig:cpu_arch}.
+
+The use of the Single Instruction Multiple Data (SIMD) paradigm enables increased
+performance on these devices by increasing data throughput for data parallel
+applications.
 
 \subsection{GPU Architecture}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
-Blah blah blah see figure \ref{fig:cpu_arch}.
+The GPU almost acts a foil to the CPU architecture.  Instead of optimizing the
+architecture of the device to increase performance of a single thread, the GPU
+architecture is specifically designed for large throughput performance.
+
+The memory hierarchy for the GPU architecture is given by Figure \ref{fig:gpu_arch}.
 

### 2011-11-29 paper/paper.tex
Added basic APU section to paper

@@
 \subsection{APU Architecture}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+The APU is a relatively new, novel architecture that attempts to 
+improve on the the discrete CPU+GPU system.  The CPU die is split
+between the typical CPU processor as well as allocating space
+for a SIMD engine, the basis for a GPU processor.  The APU has a 
+shared memory between the CPU and SIMD engine, in which accesses
+are facilitated through a high-speed memory controller.  The AMD
+Fusion technology is an example of this type of architecture~\cite{AMDFusion}.
+
+One of the main benefits of the APU is the elimination of the PCI-e
+bus, which allows for much faster transfer between CPU and GPU.  Daga et al.
+show that the AMD Fusion architecture allows for much faster data transfers
+between CPU and GPU, allowing for an overall improvement in runtime, despite
+reduced compute power on the GPU~\cite{Daga2011}.
 

### 2011-11-29 paper/paper.tex
Finished draft of introduction, and added a few sources

--- 2011-11-29 paper/paper.tex
@@
 increasingly popular for handling computations typically performed by the
-CPU.  However, many GPGPU algorithms place the entire workload on the GPU,
-leaving the CPU idle for the duration of the computation.  This paper
-investigates the effect various load-balancing schemes across the CPU and
-GPU have on overall application performance.  Additionally, the discrete and
+CPU.  However, many GPGPU algorithms place the entire workload on the Graphics
+processing unit (GPU) leaving the CPU idle for the duration of the computation.
+This paper investigates the effect various load-balancing schemes across the CPU
+and GPU have on overall application performance.  Additionally, the discrete and
 fused GPU architectures, and their impact on performance, will be investigated.
@@
 \IEEEPARstart{L}{everaging} the maximum performance from specialized hardware is one
-of the major challenges computer scientists face today.  The unique hardware architecture
-of the Graphics Processing Unit (GPU) gives it a place as a massively parallel hardware
-accelerator.  However, because of the difficulty in programming for this hardware,
-most of the effort is spent on trying to achieve performance from the GPU, while
-leaving the CPU unused.  By also using the CPU to process a smaller part of the
-data we can achieve higher utilization of resources and therefore increase program
-performance.
+of the major problems facing computer scientists today.  The unique architecture of the
+GPU gives it a place as a massively parallel hardware accelerator.  Programming frameworks
+such as OpenCL allow programmers to utilize the stream processors of the GPU for non-graphics
+data.  However, because of the difficulty in programming for this hardware, most of the
+effort is spent on trying to achieve performance from the GPU, while the CPU is left
+unused.  By also making use of CPU to process a chunk of the data, we can achieve higher
+utilization of resources and therefore increase overall program performance.
 
-GPU+CPU co-processing hardly a new idea.  Jimenez et al. \cite{jimenez} present a
+GPU+CPU co-processing hardly a new idea.  Jimenez et al.~\cite{Jimenez2009} present a
 dynamic library which can be used in a system to dynamically schedule tasks for GPU
-computation across various processes. The method describes only looks at optimizing
-usage of the GPU across the whole system, instead of optimizing a single application’s
+computation across various processes. The method described only looks at optimizing
+usage of the GPU across the whole system, rather than optimizing a single application’s
 performance. The StarPU system, on the other hand, presents a method of co-processing
-and load balancing on a per application basis \cite{augonnet}. They show that by using an
-appropriate model for the load balancing, they are able to achieve improvements in
-speed for applications which call for code acceleration multiple times.  We intend
+and load balancing on a per application basis~\cite{Augonnet2009}. They show that by
+using an appropriate model for the load balancing, they are able to achieve improvements
+in speed for applications which call for code acceleration multiple times.  We intend
 to extend this work to improve the speed of each kernel execution by splitting the
-work of that kernel over the capable resources, similar to the schemes used by OpenMP.
-Both of these papers address data transfer times as a major cost of GPU computing.
-Using the new AMD Fusion architecture, which fuses the CPU and GPU onto a single die,
-we may be able to eliminate or greatly reduce this cost, which will impact the overhead
-of GPU+CPU co-processing.
+work of that kernel over the capable resources, similar to the schemes used by
+OpenMP~\cite{Dagum1998}. Both of these papers address data transfer times as a major cost
+of GPU computing. Using the new AMD Fusion architecture, which fuses the CPU and GPU onto
+a single die, we may be able to eliminate or greatly reduce this cost, which will impact
+the overhead of GPU+CPU co-processing.
 
 We expect that by using GPU+CPU processing we will improve the performance of the
-application, when compared to CPU-only or GPU-only.  We will measure performance of the
-application using a wall time clock.  We also hypothesize that the fused GPU+CPU
-architecture will allow for lower overhead for load balancing.
+application, when compared to CPU-only or GPU-only.  That is, consider a workload of size
+$n$, a GPU that requires $t_g$ time to perform one block of computation, and a CPU that
+requires $t_c$ time to perform one block of computation. We expect that the GPU would
+require time $nt_g$ to perform a given computation, while the CPU would require $nt_c$
+time to perform the same computation.  Optimally, we would expect the load-balancing
+scheme to require the following time to perform the computation.
+\[\max{\{(n-m)t_g, mt_c}\} + t_d + t_o\]
+Where $m$ is the amount of the workload given to the CPU ($m \leq n$), $t_d$ is the data
+transfer cost, and $t_o$ is any overhead incurrent from synchronization costs of the load
+balancing scheme or otherwise.  The above will be faster than just the GPU iff
+$t_g - \max{\{(n-m)t_g, mt_c}\} < t_d + t_o$.
 
-\section{Motivation}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+Two different load balancing schemes were implemented for this paper.  The first was a
+static load balancer, which partitions the workload in two, and transfers one chunk to
+the CPU, and the other to the GPU.  The CPU and GPU work entirely independently, and
+their results are merged once each finish.  The static load balancer should have no
+synchronization overhead, and transfers all of the data prior to the computation.
+The main caveat with a static load balancer is that the ratio of the workload given to
+the CPU is chosen by the load balancer.  An optimal ratio would be one such that the
+time to perform the computation is equal for both the CPU and GPU.  Unfortunately, this
+optimal ratio will be application-dependent, meaning that a fixed ratio cannot be optimal
+for every application.  In order for a static load balancer to be optimal for a given
+application, a programmer must tune the amount of workload given to the CPU that works
+best for that application.
+
+The second scheme is a dynamic load balancer, which partitions the data into many smaller
+chunks.  One chunk is sent to the CPU and GPU each for computation.  When one finished,
+it requests another chunk from the load balancer, and this process repeats until the
+entirely workload has been sent to either the CPU or GPU.  The dynamic load balancer
+will incur some amount of synchronization overhead, because shared variables that track
+which chunks of the workload are left must be protected by a lock.  Unlike the static
+load balancer, the dynamic load balancer transfers data throughout the computation.
+The dynamic load balancer is designed to mitigate the primary problem with the static
+load balancer - that the optimal ratio of computation performed on the CPU is
+application-dependent.  A chunk size must be chosen such that the processing units are
+not requesting chunks so frequently that contention for shared variables rise and
+synchronization overhead increases, yet small enough to not allow one processing unit
+to receive a disproportionate amount of work.
 
@@
 cache pollution, thus avoiding latency.  An example CPU architecture is given
-by Figure \ref{fig:cpu_arch}.
+by Figure~\ref{fig:cpu_arch}.
 
@@
 
-The memory hierarchy for the GPU architecture is given by Figure \ref{fig:gpu_arch}.
+The memory hierarchy for the GPU architecture is given by Figure~\ref{fig:gpu_arch}.
 
@@
 \subsection{APU Architecture}
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\section{Results}
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\section{Analysis}
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
@@
 
-\bibitem{augonnet}
+\bibitem{Augonnet2009}
 C.~Augonnet, S.~Thibault, R.~Namyst, and P.A.~Wacrenier,
@@
 
-\bibitem{che}
+\bibitem{Che2009}
 S.~Che, M.~Boyer, J.~Meng, D.~Tarjan, J.W.~Sheaffer, S.H.~Lee, and K.~Skadron,
@@
 
-\bibitem{danalis}
+\bibitem{Dagum1998}
+L.~Dagum and R.~Menon,
+``OpenMP: An industry-standard API for shared-memory programming'', in
+\emph{IEEE Computational Science and Engineering} 1998, pp. 46–55.
+
+\bibitem{Danalis2010}
 A.~Danalis, G.~Marin, C.~McCurdy, J.S.~Meredith, P.C.~Roth, K.~Spafford, V.~Tipparaju, and J.S.~Vetter,
@@
 
-\bibitem{jimenez}
+\bibitem{Jimenez2009}
 V.J.~Jiménez, L.~Villanova, I.~Gelado, M.~Gil, G.~Fursin, and N.~Navarro,
 ``Predictive Runtime Code Scheduling for Heterogeneous Architectures,'' in
-\emph{Proceedings of the 4th International Conference on High Performance Embedded
-Architectures and Compilers}, 2009, pp. 19-33.
+\emph{Proceedings of the 4th International Conference on High Performance Embedded Architectures and Compilers}, 2009, pp. 19-33.
+
+\bibitem{Ryoo2007}
+S.~Ryoo, C.I.~Rodrigues, S.S.~Baghsorkhi, S.S.~Stone, D.B.~Kirk, and W.W.~Hwu,
+``Optimization Principles and Application Performance Evaluation of a Multithreaded GPU Using CUDA'' in
+\emph{Proceedings of the 13th ACM SIGPLAN Symposium on Principles and practice of parallel programming}, 2007, pp. 73-82.
+
+\bibitem{Verner2011}
+U.~Verner, A.~Schuster, and M.~Silberstein,
+``Processing data streams with hard real-time constraints on heterogeneous systems'' in
+\emph{Proceedings of the international conference on Supercomputing}, 2011, pp. 120-130.
 

### 2011-11-29 paper/paper.tex
Added a few more references.

@@
 
+\bibitem{Daga2011}
+M.~Daga, A.M.~Aji, and W.~Feng,
+``On the Efficacy of a Fused CPU+GPU Processor (or APU) for Parallel Computing'' in
+\emph{Application Accelerators in High-Performance Computing (SAAHPC)}, 2011, pp. 141-149.
+
 \bibitem{Dagum1998}
 L.~Dagum and R.~Menon,
-``OpenMP: An industry-standard API for shared-memory programming'', in
+``OpenMP: An industry-standard API for shared-memory programming'' in
 \emph{IEEE Computational Science and Engineering} 1998, pp. 46–55.
@@
 
+\bibitem{Jablin2011}
+J.A.~Jablin, P.~McCormick, and M.~Herlihy,
+``Scout: High-Performance Heterogeneous Computing Made Simple'' in
+\emph{Parallel and Distributed Processing Workshops and Phd Forum (IPDPSW)}, 2011, pp. 2093-2096.
+
 \bibitem{Jimenez2009}
 V.J.~Jiménez, L.~Villanova, I.~Gelado, M.~Gil, G.~Fursin, and N.~Navarro,
-``Predictive Runtime Code Scheduling for Heterogeneous Architectures,'' in
+``Predictive Runtime Code Scheduling for Heterogeneous Architectures'' in
 \emph{Proceedings of the 4th International Conference on High Performance Embedded Architectures and Compilers}, 2009, pp. 19-33.

### 2011-11-30 VectorAdd.c
Fixed CPUBound kernel to match VectorAdd.  Fixed timing code in VectorAdd

@@ -24,7 +24,7 @@ struct timespec timer1;
 struct timespec timer2;
 
 //OpenCL Constructs
-const char *KernelSourceFile = "VectorAdd.cl";
+const char *KernelSourceFile = "CPUBound.cl";
 cl_platform_id platform_id;
 cl_device_id device_id_gpu;
 cl_device_id device_id_cpu;
@@ -111,7 +111,7 @@ cl_kernel create_kernel(const char* filename, const char* kernel, const cl_conte
 	cl_program program = createProgramFromSource(filename, context);
 
 	// Build the program executable
-	int err = clBuildProgram(program, 1, &device, NULL, NULL, NULL);
+	int err = clBuildProgram(program, 1, &device, "-cl-opt-disable", NULL, NULL);
 	if (err == CL_BUILD_PROGRAM_FAILURE)
 	{
 		char *log;
@@ -261,6 +261,8 @@ void test_init()
 
 void test_chunk_setup(cl_context context, cl_command_queue queue, size_t size, size_t offset, int isGPU)
 {
+	if(size == 0)
+		return;
 	cl_mem* d_a = isGPU ? &dg_a : &dc_a;
 	cl_mem* d_b = isGPU ? &dg_b : &dc_b;
 	cl_mem* d_c = isGPU ? &dg_c : &dc_c;
@@ -295,6 +297,8 @@ void test_chunk_setup(cl_context context, cl_command_queue queue, size_t size, s
 
 void test_chunk_kernel(cl_context context, cl_command_queue queue, cl_device_id device, cl_kernel kernel, size_t size, size_t offset, int isGPU)
 {
+	if(size == 0)
+		return;
 	cl_mem* d_a = isGPU ? &dg_a : &dc_a;
 	cl_mem* d_b = isGPU ? &dg_b : &dc_b;
 	cl_mem* d_c = isGPU ? &dg_c : &dc_c;
@@ -315,6 +319,8 @@ void test_chunk_kernel(cl_context context, cl_command_queue queue, cl_device_id 
 
 void test_chunk_cleanup(cl_context context, cl_command_queue queue, size_t size, size_t offset, int isGPU)
 {
+	if(size == 0)
+		return;
 	cl_mem* d_a = isGPU ? &dg_a : &dc_a;
 	cl_mem* d_b = isGPU ? &dg_b : &dc_b;
 	cl_mem* d_c = isGPU ? &dg_c : &dc_c;
[…]

### 2011-11-30 VectorAdd.c
Fixed static scheduling to give better than GPU-only performance

@@ -3,6 +3,7 @@
 #include <string.h>
 #include <time.h>
 #include <pthread.h>
+#include <unistd.h>
 
 #ifdef __APPLE__
 #include <OpenCL/opencl.h>
@@ -24,7 +25,7 @@ struct timespec timer1;
 struct timespec timer2;
 
 //OpenCL Constructs
-const char *KernelSourceFile = "CPUBound.cl";
+const char *KernelSourceFile = "VectorAdd.cl";
 cl_platform_id platform_id;
 cl_device_id device_id_gpu;
 cl_device_id device_id_cpu;
@@ -36,8 +37,11 @@ cl_program program;
 cl_kernel kernel_compute_cpu;
 cl_kernel kernel_compute_gpu;
 
+cl_event event_gpu;
+cl_event event_cpu;
+
 //Number of iterations to warmup caches
-const int warmup = 2;
+const int warmup = 0;
 
 enum scheme_t { CPU_ONLY, GPU_ONLY, CPU_GPU_STATIC, CPU_GPU_DYNAMIC };
 enum scheme_t scheme = CPU_ONLY;
@@ -111,7 +115,8 @@ cl_kernel create_kernel(const char* filename, const char* kernel, const cl_conte
 	cl_program program = createProgramFromSource(filename, context);
 
 	// Build the program executable
-	int err = clBuildProgram(program, 1, &device, "-cl-opt-disable", NULL, NULL);
+	//int err = clBuildProgram(program, 1, &device, "-cl-opt-disable", NULL, NULL);
+	int err = clBuildProgram(program, 1, &device, NULL, NULL, NULL);
 	if (err == CL_BUILD_PROGRAM_FAILURE)
 	{
 		char *log;
@@ -303,6 +308,8 @@ void test_chunk_kernel(cl_context context, cl_command_queue queue, cl_device_id 
 	cl_mem* d_b = isGPU ? &dg_b : &dc_b;
 	cl_mem* d_c = isGPU ? &dg_c : &dc_c;
 
+	cl_event* event = isGPU ? &event_gpu : &event_cpu;
+
 	size_t local_size;
 	clGetDeviceInfo(device, CL_DEVICE_MAX_WORK_GROUP_SIZE, sizeof(size_t), &local_size, NULL);
 
@@ -313,7 +320,7 @@ void test_chunk_kernel(cl_context context, cl_command_queue queue, cl_device_id 
 	CHKERR(err, "Errors setting kernel arguments");
 
[…]

### 2011-12-01 VectorAdd.c
Added end-to-end timing

@@ -24,6 +24,12 @@
 struct timespec timer1;
 struct timespec timer2;
 
+#define TOTAL_TIMER_START clock_gettime(CLOCK_REALTIME, &total_timer1)
+#define TOTAL_TIMER_END clock_gettime(CLOCK_REALTIME, &total_timer2)
+#define TOTAL_MILLISECONDS (total_timer2.tv_sec - total_timer1.tv_sec) * 1000.0f + (total_timer2.tv_nsec - total_timer1.tv_nsec) / 1000000.0f
+struct timespec total_timer1;
+struct timespec total_timer2;
+
 //OpenCL Constructs
 const char *KernelSourceFile = "VectorAdd.cl";
 cl_platform_id platform_id;
@@ -345,9 +351,10 @@ void test_cleanup()
 	//verify_answer(h_c, h_check, length);
 }
 
-void run_test(float* data_time, float* exec_time)
+void run_test(float* data_time, float* exec_time, float* total_time)
 {
 	test_setup();
+	TOTAL_TIMER_START;
 	TIMER_START;
 	test_init();	
 	TIMER_END;
@@ -451,6 +458,8 @@ void run_test(float* data_time, float* exec_time)
 		abort();
 	}
 	test_cleanup();	
+	TOTAL_TIMER_END;
+	*total_time = TOTAL_MILLISECONDS;
 }
 
 void fillArray(unsigned char* nums, unsigned long length)
@@ -518,6 +527,7 @@ int main(int argc, char** argv)
 
 	float data_time = 0;
 	float exec_time = 0;
+	float total_time = 0;
 	
 	int i;
 	for(i = 0; i < iters+warmup; i++)
@@ -525,10 +535,10 @@ int main(int argc, char** argv)
 		memset(h_c, 0, sizeof(unsigned char) * length);
 //		vadd_default(h_a, h_b, h_c, length, &data_time, &exec_time);
 //		verify_answer(h_c, h_check, length);	
-		run_test(&data_time, &exec_time);
+		run_test(&data_time, &exec_time, &total_time);
 		if(i >= warmup)
 		{
-			fprintf(stdout,"%d\tVectorAdd\t%s\t%f\t%lu\t%f\t%f\n", i - warmup, scheme_name, ratio, length, data_time, exec_time);
[…]

### 2011-12-02 Reduce.c
Wrote Reduce kernel and application.

@@ -1,0 +1,550 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <time.h>
+#include <pthread.h>
+#include <unistd.h>
+
+#ifdef __APPLE__
+#include <OpenCL/opencl.h>
+#else
+#include <CL/opencl.h>
+#endif
+
+#define CHKERR(err, str) \
+	if (err != CL_SUCCESS) \
+	{ \
+		fprintf(stdout, "CL Error %d: %s\n", err, str); \
+		exit(1); \
+	}
+
+#define TIMER_START clock_gettime(CLOCK_REALTIME, &timer1)
+#define TIMER_END clock_gettime(CLOCK_REALTIME, &timer2)
+#define MILLISECONDS (timer2.tv_sec - timer1.tv_sec) * 1000.0f + (timer2.tv_nsec - timer1.tv_nsec) / 1000000.0f
+struct timespec timer1;
+struct timespec timer2;
+
+#define TOTAL_TIMER_START clock_gettime(CLOCK_REALTIME, &total_timer1)
+#define TOTAL_TIMER_END clock_gettime(CLOCK_REALTIME, &total_timer2)
+#define TOTAL_MILLISECONDS (total_timer2.tv_sec - total_timer1.tv_sec) * 1000.0f + (total_timer2.tv_nsec - total_timer1.tv_nsec) / 1000000.0f
+struct timespec total_timer1;
+struct timespec total_timer2;
+
+typedef unsigned long reduce_t;
+
+//OpenCL Constructs
+const char *KernelSourceFile_cpu = "Reduction_CPU.cl";
+const char *KernelSourceFile_gpu = "Reduction_GPU.cl";
+cl_platform_id platform_id;
+cl_device_id device_id_gpu;
+cl_device_id device_id_cpu;
+cl_context context_cpu;
+cl_context context_gpu;
+cl_command_queue commands_cpu;
+cl_command_queue commands_gpu;
+cl_program program;
+cl_kernel kernel_compute_cpu;
+cl_kernel kernel_compute_gpu;
+
+cl_event event_gpu;
+cl_event event_cpu;
+
+//Number of iterations to warmup caches
+const int warmup = 2;
+
+enum scheme_t { CPU_ONLY, GPU_ONLY, CPU_GPU_STATIC, CPU_GPU_DYNAMIC };
+enum scheme_t scheme = CPU_ONLY;
[…]

### 2011-12-02 Reduce.c
Added optimziations for GPU reduce

@@ -264,6 +264,8 @@ void* dynamic_scheduler(void* argv)
 
 void test_setup()
 {
+	ans_gpu = 0;
+	ans_cpu = 0;
 	fillArray(h_a, length);
 	serial_reduce(h_a, &h_check, length);
 }
@@ -305,7 +307,7 @@ void test_chunk_kernel(cl_context context, cl_command_queue queue, cl_device_id 
 		return;
 	cl_mem* d_a = isGPU ? &dg_a : &dc_a;
 	cl_mem* d_b = isGPU ? &dg_b : &dc_b;
-	size_t chunk = isGPU ? 1 : size;
+	size_t chunk = isGPU ? 2 : size;
 
 	cl_event* event = isGPU ? &event_gpu : &event_cpu;
 
@@ -319,7 +321,7 @@ void test_chunk_kernel(cl_context context, cl_command_queue queue, cl_device_id 
 	err |= clSetKernelArg(kernel, 2, sizeof(size_t), &size);
 	err |= clSetKernelArg(kernel, 3, sizeof(size_t), &chunk);
 	if(isGPU)
-		err |= clSetKernelArg(kernel, 4, sizeof(reduce_t) * local_size, NULL);
+		err |= clSetKernelArg(kernel, 4, sizeof(reduce_t) * local_size * 2, NULL);
 	CHKERR(err, "Errors setting kernel arguments");
 
 	size_t groups = size / local_size / chunk + (size % (local_size*chunk) == 0 ? 0 : 1);
@@ -423,8 +425,8 @@ void run_test(float* data_time, float* exec_time, float* total_time)
 		test_chunk_kernel(context_cpu, commands_cpu, device_id_cpu, kernel_compute_cpu, length - gpu_size, 0, 0);
 		clFlush(commands_gpu);
 		clFlush(commands_cpu);
-		cl_event events[2] = {event_cpu, event_gpu};
-		clEnqueueWaitForEvents(commands_cpu, 2, events);
+		//cl_event events[2] = {event_cpu, event_gpu};
+		//clEnqueueWaitForEvents(commands_cpu, 2, events);
 		clFinish(commands_cpu);
 		clFinish(commands_gpu);
 		TIMER_END;

### 2011-12-02 VectorAddPlus.c
Added VectorAddPlus for CPU-boundedness

@@ -30,11 +30,8 @@ struct timespec timer2;
 struct timespec total_timer1;
 struct timespec total_timer2;
 
-typedef unsigned long reduce_t;
-
 //OpenCL Constructs
-const char *KernelSourceFile_cpu = "Reduction_CPU.cl";
-const char *KernelSourceFile_gpu = "Reduction_GPU.cl";
+const char *KernelSourceFile = "CPUBound.cl";
 cl_platform_id platform_id;
 cl_device_id device_id_gpu;
 cl_device_id device_id_cpu;
@@ -50,7 +47,7 @@ cl_event event_gpu;
 cl_event event_cpu;
 
 //Number of iterations to warmup caches
-const int warmup = 2;
+const int warmup = 0;
 
 enum scheme_t { CPU_ONLY, GPU_ONLY, CPU_GPU_STATIC, CPU_GPU_DYNAMIC };
 enum scheme_t scheme = CPU_ONLY;
@@ -58,16 +55,17 @@ float ratio = 0.01;
 
 //Data
 unsigned long length;
-reduce_t* h_a;
-reduce_t* h_b;
-reduce_t h_check;
+unsigned char* h_a;
+unsigned char* h_b;
+unsigned char* h_c;
+unsigned char* h_check;
 cl_mem dc_a;
 cl_mem dc_b;
+cl_mem dc_c;
 cl_mem dg_a;
 cl_mem dg_b;
-reduce_t ans_gpu = 0;
-reduce_t ans_cpu = 0;
-reduce_t ans;
+cl_mem dg_c;
+
 
 // Struct for passing arguments to dynamic_scheduler
 struct dynamic_args
@@ -77,11 +75,10 @@ struct dynamic_args
 	float exec_time;
 };
 
-
 //Function Prototypes
-void fillArray(reduce_t* nums, const unsigned long length);
-void verify_answer(reduce_t* toCheck, reduce_t* answer, const unsigned int len);
-void serial_reduce(reduce_t* a, reduce_t* c, const unsigned int len);
+void fillArray(unsigned char* nums, unsigned long length);
+void verify_answer(unsigned char* toCheck, unsigned char* answer, const unsigned int len);
+void serial_vector_add(unsigned char* a, unsigned char* b, unsigned char* c, const unsigned int len);
 
 void test_setup();
 void test_init();
[…]

### 2011-12-04 paper/paper.tex
Attempted to single-column appendix and a few other things.
Added CPU bound and CPU reduction benchmarks.
Intermittent commit because I'm paranoid.

@@
 \usepackage[cmex10]{amsmath}
+\usepackage{ifthen}
+\usepackage{stfloats}
 
@@
 
-\begin{figure}[!t]
+\begin{figure}[t]
 \centering
@@
 
+\section{Load Balancing Benchmarks}
+
+For this paper, three different benchmarks were used to assess the
+effectiveness of the load balancer.  The first is a reduction benchmark,
+in which the sum of variably-sized array is computed.  The reduction benchmark
+is unique in that different code is being run on the CPU and the GPU.  The
+reason for this is that the GPU cannot efficiently compute a reduction the way
+a CPU typically would -- due to the relative innefficiency of backward branches
+on the GPU.  A reduction on the GPU is a multi-stage process, in which the
+output of one step is used as the input of the next.  Each steam processor
+computes the reduction of two elements during each iteration, reducing the amount
+of work during each step by a factor of two.  As a result, each step utilizes half
+the stream processors of the previous step.  This continues until the reduction
+operation is only done by a single stream processor.  A natural consequence of
+this algorithm is that not all of the stream processors will be utilized during
+the computation.
+
+The next benchmark is a simple vector addition operation that computes the sum
+of two input vectors, and stores the result into an output vector.  The third
+benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
+the Vector Add benchmark, but it computes 1000 additions before storing the result
+in the output vector.  The motivation behind this benchmark is to reduce memory
+access overhead with more computation.
+
 \section{Results}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{static_discrete}
+\caption{Varying Data Ratio with a Discrete GPU}
+\label{fig:static_discrete}
+\end{figure}
+
+Consider Figure \ref{fig:static_discrete}, where the amount of data
+executed on the GPU is varied between 90\% and 100\%.  The total execution
+time including data transfers has an obvious downward trend as more
+data is executed on the GPU.  However, there are a couple of static
+ratios which yielded better performance than executing all of the data
+on the GPU.  Most notably, when only 98.6\% of the data was executed on
+the GPU.  In this instance, the overall execution time was 325.27 seconds,
+compared to 333.31 seconds with only the GPU -- a ~2.4\% speedup.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{reduce_fused}
+\caption{Reduction Benchmark on an APU Architecture}
+\label{fig:reduce_fused}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_fused}
+\caption{Vector Add Benchmark on an APU Architecture}
+\label{fig:vector_fused}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_plus_fused}
+\caption{Vector Add+ Benchmark on an APU Architecture}
+\label{fig:vector_plus_fused}
+\end{figure}
+
+Next, consider Figures \ref{fig:reduce_fused}, \ref{fig:vector_fused}, and
+\ref{fig:vector_plus_fused}. The figures demonstrate a variety of different
+load balancing options for the Reduction, Vector Add, and Vector Add+
+benchmarks, across data sizes ranging between 1 MB and 16 MB.  Each figure
+depicts the total execution time for the CPU-only and GPU-only computations,
+as well as static load balancing with 25\%, 50\%, and 75\% of the data on the
+GPU.  In each of the figures, it's evident that the GPU performs significantly
+faster than the CPU for the given benchmarks, with the static partitioning
+schemes lying somewhere between the two.
 
@@
 
+\setboolean{@twocolumn}{false}
 \newpage

### 2011-12-04 paper/paper.tex
Just some intermittent changes.

@@
 The dynamic load balancer is designed to mitigate the primary problem with the static
-load balancer - that the optimal ratio of computation performed on the CPU is
+load balancer -- that the optimal ratio of computation performed on the CPU is
 application-dependent.  A chunk size must be chosen such that the processing units are
@@
 
-\section{Analysis}
+The only deviation from the above is the reduction benchmark, in which the GPU
+only performs ~14\% faster than the CPU in the 16 MB case (0.104 seconds for the
+GPU, 0.091 seconds for the CPU), and some of the static load balancing schemes
+perform slightly better than just the GPU.  For instance, in the 9 MB and 10 MB
+cases, both the 50\% and 75\% static schemes performed better than just the GPU
+(0.5\% and 4.4\% respective speedups for 9 MB, 1.5\% and 3.6\% respective speedups
+for 10 MB).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{reduce_static_fused}
+\caption{Reduction Benchmark with Varied Execution Ratio}
+\label{fig:reduce_static_fused}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_static_fused}
+\caption{Vector Add Benchmark with Varied Execution Ratio}
+\label{fig:vector_static_fused}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_plus_static_fused}
+\caption{Vector Add+ Benchmark with Varied Execution Ratio}
+\label{fig:vector_plus_static_fused}
+\end{figure}
+
+Finally, consider Figures \ref{fig:reduce_static_fused}, \ref{fig:vector_static_fused},
+and \ref{fig:vector_plus_static_fused}. In each, the runtim for both the CPU and GPU are
+shown with a 16 MB problem size.  Additionally, a variety of static execution ratios are
+shown between 0.1\% and 99.9\%.  In every benchmark, none of the execution ratios were
+able to achieve a better time than just the GPU.
+
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
@@
 sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
+eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
+enim ad minim veniam, quis nostrud exercitation ullamco laboris
+nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
+in reprehenderit in voluptate velit esse cillum dolore eu fugiat
+nulla pariatur. Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\section{Analysis}
+Based on the times shown in the previous figures, it is clear that the
+static load balancing is ineffective at improving the overal efficiency
+of any of the benchmarks tested.  There were a few scenarios in which
+a specific static execution ratio achieved meager speedup (<5\%) over
+just the GPU, but these situations were scarse, and did not achieve the
+speedup we desired.  Additionally, this small amount of speedup would
+require a significant amount of tine-tuning by a programmer.
+
+Consider the 16 MB vector add benchmark, where the CPU-only case took
+0.07465 seconds to run, and the GPU-only case took 0.02873 second to run.
+Ideally, 72.2\% of the workload would be placed on the GPU, and 27.8\%
+on the CPU.  Assuming no other overhead, this would result in an overall
+runtime of 0.02075 seconds -- a 27.8\% speedup over the GPU.  Using similar
+reasoning, a 43.6\% theoretical speedup is attainable for the reduction
+benchmark.  A measly 2\% would be expected from the vector add+ benchmark,
+because of the vast disparity in the CPU-only and GPU-only runtimes.
 

### 2011-12-05 paper/paper.tex
Merged the Results and Analysis sections, and included the discrete GPU results.

--- 2011-12-04 paper/paper.tex
@@
 in the output vector.  The motivation behind this benchmark is to reduce memory
-access overhead with more computation.
+access overhead with more computation.  All numbers reported in the results section
+are given as a mean of ten runs for that particular benchmark, load-balancing scheme,
+and data size.
 
-\section{Results}
+\section{Results and Analysis}
+The three implemented benchmarks were run on a system with a discrete GPU, as well as
+on a fused GPU platform.  In dynamic load balancing benchmarks, a chunk size of 80~KB
+was used.  Dynamic results are only given for the discrete GPU testing environment,
+as the APU platform was running Windows and our code was written with POSIX compliance
+in mind.  In static load balancing benchmarks, the ratio of data computed on the GPU
+will be specified.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{reduce_discrete}
+\caption{Reduction Benchmark on a Discrete GPU Platform}
+\label{fig:reduce_discrete}
+\end{figure}
+
+First, the reduction benchmark was run on the discrete GPU platform with a variety of
+different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
+Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
+due to the GPU not utilizing all of its stream processors during a reduction operation, as
+discussed previously.  Unfortunately, none of the static load balancing schemes were able
+to outperform just the CPU; all the static times were somewhere inbetween the CPU-only and
+GPU-only times.  Additionally, the dynamic load balancing scheme performed significantly
+worse than all the other schemes, and more than 100\% worse than just the CPU.
+100\% worse than just the CPU.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_discrete}
+\caption{Vector Add Benchmark on a Discrete GPU Platform}
+\label{fig:vector_discrete}
+\end{figure}
+
+Now, consider Figure \ref{fig:vector_discrete}, where the vector add benchmark was run on
+the discrete GPU platform.  In this case, we see that the GPU has nearly a 4x speedup over
+just the CPU.  However, similar to the reduction benchmark, the static load balancing schemes
+never outperform just the GPU alone, with all of them lying somewhere between the CPU-only
+and GPU-only runs.  Also similar to the reduction benchmark, the dynamic load balanching scheme
+performs significantly worse than all the other schemes.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_plus_discrete}
+\caption{Vector Add+ Benchmark on a Discrete GPU Platform}
+\label{fig:vector_plus_discrete}
+\end{figure}
+
+The results for the vector add+ benchmark on the discrete GPU platform are shown in Figure
+\ref{fig:vector_plus_discrete}.  There are a few notable trends in this figure that are not
+present in the previous two.  The first is is how drastically the GPU outperforms the CPU --
+with over a 100x speedup over just the CPU with 16~MB of data.  The second is that the
+dynamic load balancing drastically outperforms the static load balancing schemes.  This is
+almost certainly because the GPU so drastically outperforms the CPU.  With the dynamic load
+balancing scheme, only small chunks of data are given to the CPU at a time, rather than a
+static percentage.
+
 \begin{figure}[t]
@@
 
-Consider Figure \ref{fig:static_discrete}, where the amount of data
-executed on the GPU is varied between 90\% and 100\%.  The total execution
-time including data transfers has an obvious downward trend as more
-data is executed on the GPU.  However, there are a couple of static
-ratios which yielded better performance than executing all of the data
-on the GPU.  Most notably, when only 98.6\% of the data was executed on
-the GPU.  In this instance, the overall execution time was 325.27 seconds,
-compared to 333.31 seconds with only the GPU -- a ~2.4\% speedup.
+For the above figures, the static load balancer was run with three different
+data ratios run on the GPU -- 25\%, 50\%, and 75\%.  The general trend was that
+as more of the workload was placed on the GPU, the runtime improved.  As a
+result, we run the vector add benchmark again with a high granularity of data
+ratios.  Figure \ref{fig:static_discrete} shows  this benchmark on the discrete
+GPU platform with between 90\% and 100\% of the data on the GPU, with a 0.1\%
+granularity.  The total execution time, including data transfers, has an
+obvious downward trend as more of the data is executed on the GPU.  That said,
+there are a couple of static ratios which yielded better performance than
+executing all of the data on the GPU.  Most notably, when only 98.6\% of the
+data was executed on the GPU.  In this instance, the overall execution time was
+325.27 seconds, compared to 333.31 seconds with only the GPU -- a ~2.4\%
+speedup.  However, these few instances of speedup may simply be due to variability.
 
@@
 \includegraphics[width=3.0in]{reduce_fused}
-\caption{Reduction Benchmark on an APU Architecture}
+\caption{Reduction Benchmark on a Fused CPU+GPU Platform}
 \label{fig:reduce_fused}
@@
 \includegraphics[width=3.0in]{vector_fused}
-\caption{Vector Add Benchmark on an APU Architecture}
+\caption{Vector Add Benchmark on a Fused CPU+GPU Platform}
 \label{fig:vector_fused}
@@
 \includegraphics[width=3.0in]{vector_plus_fused}
-\caption{Vector Add+ Benchmark on an APU Architecture}
+\caption{Vector Add+ Benchmark on a Fused CPU+GPU Platform}
 \label{fig:vector_plus_fused}
@@
 load balancing options for the Reduction, Vector Add, and Vector Add+
-benchmarks, across data sizes ranging between 1 MB and 16 MB.  Each figure
+benchmarks, across data sizes ranging between 1~MB and 16~MB.  Each figure
 depicts the total execution time for the CPU-only and GPU-only computations,
@@
 
-\section{Analysis}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+The only deviation from the above is the reduction benchmark, in which the GPU
+only performs ~14\% faster than the CPU in the 16~MB case (0.104 seconds for the
+GPU, 0.091 seconds for the CPU), and some of the static load balancing schemes
+perform slightly better than just the GPU.  For instance, in the 9~MB and 10~MB
+cases, both the 50\% and 75\% static schemes performed better than just the GPU
+(0.5\% and 4.4\% respective speedups for 9~MB, 1.5\% and 3.6\% respective speedups
+for 10~MB). What is also interested to note is that on the fused CPU+GPU architecture,
+the GPU outperformed the CPU on the reduction benchmark.  However, the opposite was
+true of the reduction benchmark on the discrete GPU platform.
+
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=3.0in]{reduce_static_fused}
+%\caption{Reduction Benchmark with Varied Execution Ratio}
+%\label{fig:reduce_static_fused}
+%\end{figure}
+
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=3.0in]{vector_static_fused}
+%\caption{Vector Add Benchmark with Varied Execution Ratio}
+%\label{fig:vector_static_fused}
+%\end{figure}
+
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=3.0in]{vector_plus_static_fused}
+%\caption{Vector Add+ Benchmark with Varied Execution Ratio}
+%\label{fig:vector_plus_static_fused}
+%\end{figure}
+
+%Finally, consider Figures \ref{fig:reduce_static_fused}, \ref{fig:vector_static_fused},
+%and \ref{fig:vector_plus_static_fused}. In each, the runtim for both the CPU and GPU are
+%shown with a 16~MB problem size.  Additionally, a variety of static execution ratios are
+%shown between 0.1\% and 99.9\%.  In every benchmark, none of the execution ratios were
+%able to achieve a better time than just the GPU.
+
+Based on the times shown in the previous figures, it is clear that the
+static load balancing is ineffective at improving the overal efficiency
+of any of the benchmarks tested.  There were a few scenarios in which
+a specific static execution ratio achieved meager speedup (<5\%) over
+just the GPU, but these situations were scarse, and did not achieve the
+speedup we desired.  Additionally, this small amount of speedup would
+require a significant amount of tine-tuning by a programmer.
+
+Consider the 16~MB vector add benchmark, where the CPU-only case took
+0.07465 seconds to run, and the GPU-only case took 0.02873 second to run.
+Ideally, 72.2\% of the workload would be placed on the GPU, and 27.8\%
+on the CPU.  Assuming no other overhead, this would result in an overall
+runtime of 0.02075 seconds -- a 27.8\% speedup over the GPU.  Using similar
+reasoning, a 43.6\% theoretical speedup is attainable for the reduction
+benchmark.  A measly 2\% would be expected from the vector add+ benchmark,
+because of the vast disparity in the CPU-only and GPU-only runtimes.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{sequential_discrepancy}
+\caption{Discrepancies between expected runtimes and actual runtimes.}
+\label{fig:sequential_discrepancy}
+\end{figure}
+
+The above results have lead us to believe that the CPU and GPU kernels are
+not running truely in parallel as we initially believed.  As such, we used
+an AMD profiler to report how long each of the kernels took using the static
+load balancer with 95\% of the data on the CPU.  Figure \ref{fig:sequential_discrepancy}
+shows the discrepancy between the actual total runtime of the vector add
+benchmark and the expected runtime (that is, the sum of the CPU kernel runtime
+and the GPU kernel runtime).  If the kernels were truely running in parallel,
+we would expect to see a negative discrepancy between the actual runtime and
+expected runtime.
+
+Many of the trials produced a small positive discrepancy.  There were only
+two trials which produced a negative discrepancy; however, these trials
+were unusual in other ways as well.  The overall runtime of the CPU kernel
+was rough 5.1~ms in all of the trials except these two, which were 36.6~ms
+and 31.6~ms, respectively.  We believe that OpenCL is beginning the CPU
+kernel, scheduling the GPU kernel before the CPU kernel has completed,
+running the GPU kernel to completion, then scheduling the CPU kernel again
+and letting it run to completion.  Thus, the profiler's record of the runtime
+of the CPU kernel would include the runtime of the GPU kernel.
 

### 2011-12-05 paper/paper.tex
Fleshed out architecture section.

@@
 of the CPU typically only runs one or two hardware-enabled threads to reduce
-cache pollution, thus avoiding latency.  An example CPU architecture is given
+cache pollution, thus avoiding latency.  An simplistic example CPU architecture is given
 by Figure~\ref{fig:cpu_arch}.
@@
 performance on these devices by increasing data throughput for data parallel
-applications.
+applications.  However, optimal performance of these types of instructions often
+involves complex compilation techniques and hand-tuning of the assembly.
 
@@
 architecture of the device to increase performance of a single thread, the GPU
-architecture is specifically designed for large throughput performance.
+architecture is specifically designed for large throughput performance.  Large
+numbers of threads can be easily created an run on a GPU.  Large register files
+enable tens of thousands of threads to be run on today's GPU architecture with
+very little penalty for context switching.
 
-The memory hierarchy for the GPU architecture is given by Figure~\ref{fig:gpu_arch}.
+Groups of proccessing units, Streaming Multiprocessors (SMs), share a single instruction
+dispatch unit and operate in lock step.  The threads themselves run on each SM
+in a group, called a warp, typically of 32 threads.  Because of these architectual
+features divergent branching in a warp (caused by if-else clauses) greatly degrade application
+performance.
+
+The memory architecture for the GPU is very different than that found on the CPU. 
+Instead of a large hierarchy of caches to hide memory latency for single thread
+performance the latecy to memory is instead hidden through aggressive context
+switching.  Newer GPU architectures may contain primative caching schemes.
+Each SM on the GPU also contains a local memory memory space which can accessed 
+at much lower latency .  This memory space is typically used as a user-managed
+cache. The memory hierarchy for the GPU architecture is given by Figure~\ref{fig:gpu_arch}.
 

### 2011-12-05 paper/paper.tex
Added Load-balacing schemes section.

@@
 and data size.
+
+\section{Load Balancing Schemes}
+A naive implementation of CPU+GPU co-processing will in many cases lead to an imbalance
+of work for either the CPU or GPU.  To deal with this problem we will investigate the
+use of 2 different load balancing schemes in addtion to the performance of a CPU-only or
+a GPU-only solution.
+
+The first load balancing scheme that was implemented was a static division of work between
+the CPU and the GPU at a specified ratio.  This ratio was user-defined allowing for best 
+performance after runtime benchmarking.  The main cause of overhead for this method of
+load-balancing is that of a varying workload.  If the workload for the kernel is imbalanced
+between workgroups, it is impossible to statically determine the proper amount of work
+to put on each kernel.  Improper allocation of work to each device will not allow both
+devices to be fully utilized, and application performance will not be optimal.
+
+To address the issue of a dynamic work load between workgroups for kernels, we also used
+a dynamic load-balancing scheme.  In this scheme the worksgroups were grouped into larger
+chunks.  Each device is initially allocated one chunk of work to work on.  As each chunk
+is retired by the device a new chunk is allocated.  This load-balancing scheme will still
+not guarentee optimal performance.  Within a work chunk, a single long-running workgroup
+can hold up the entire device, not allowing other kernels to execute on open SMs.
 

### 2011-12-05 paper/paper.tex
Spellchecking on paper.

@@
 \begin{abstract}
-Leveraging maximimum performance from specialized hardware is one of the
+Leveraging maximum performance from specialized hardware is one of the
 major problems facing computer scientists today.  General-purpose computing
@@
 computation across various processes. The method described only looks at optimizing
-usage of the GPU across the whole system, rather than optimizing a single application’s
+usage of the GPU across the whole system, rather than optimizing a single applications
 performance. The StarPU system, on the other hand, presents a method of co-processing
@@
 
-Groups of proccessing units, Streaming Multiprocessors (SMs), share a single instruction
+Groups of processing units, Streaming Multiprocessors (SMs), share a single instruction
 dispatch unit and operate in lock step.  The threads themselves run on each SM
-in a group, called a warp, typically of 32 threads.  Because of these architectual
+in a group, called a warp, typically of 32 threads.  Because of these architectural
 features divergent branching in a warp (caused by if-else clauses) greatly degrade application
@@
 Instead of a large hierarchy of caches to hide memory latency for single thread
-performance the latecy to memory is instead hidden through aggressive context
-switching.  Newer GPU architectures may contain primative caching schemes.
+performance the latency to memory is instead hidden through aggressive context
+switching.  Newer GPU architectures may contain privative caching schemes.
 Each SM on the GPU also contains a local memory memory space which can accessed 
@@
 reason for this is that the GPU cannot efficiently compute a reduction the way
-a CPU typically would -- due to the relative innefficiency of backward branches
+a CPU typically would -- due to the relative inefficiency of backward branches
 on the GPU.  A reduction on the GPU is a multi-stage process, in which the
@@
 of work for either the CPU or GPU.  To deal with this problem we will investigate the
-use of 2 different load balancing schemes in addtion to the performance of a CPU-only or
+use of 2 different load balancing schemes in addition to the performance of a CPU-only or
 a GPU-only solution.
@@
 To address the issue of a dynamic work load between workgroups for kernels, we also used
-a dynamic load-balancing scheme.  In this scheme the worksgroups were grouped into larger
+a dynamic load-balancing scheme.  In this scheme the workgroups were grouped into larger
 chunks.  Each device is initially allocated one chunk of work to work on.  As each chunk
 is retired by the device a new chunk is allocated.  This load-balancing scheme will still
-not guarentee optimal performance.  Within a work chunk, a single long-running workgroup
+not guarantee optimal performance.  Within a work chunk, a single long-running workgroup
 can hold up the entire device, not allowing other kernels to execute on open SMs.
@@
 discussed previously.  Unfortunately, none of the static load balancing schemes were able
-to outperform just the CPU; all the static times were somewhere inbetween the CPU-only and
+to outperform just the CPU; all the static times were somewhere in between the CPU-only and
 GPU-only times.  Additionally, the dynamic load balancing scheme performed significantly
@@
 never outperform just the GPU alone, with all of them lying somewhere between the CPU-only
-and GPU-only runs.  Also similar to the reduction benchmark, the dynamic load balanching scheme
+and GPU-only runs.  Also similar to the reduction benchmark, the dynamic load balancing scheme
 performs significantly worse than all the other schemes.
@@
 Based on the times shown in the previous figures, it is clear that the
-static load balancing is ineffective at improving the overal efficiency
+static load balancing is ineffective at improving the overall efficiency
 of any of the benchmarks tested.  There were a few scenarios in which
 a specific static execution ratio achieved meager speedup (<5\%) over
-just the GPU, but these situations were scarse, and did not achieve the
+just the GPU, but these situations were scarce, and did not achieve the
 speedup we desired.  Additionally, this small amount of speedup would
@@
 The above results have lead us to believe that the CPU and GPU kernels are
-not running truely in parallel as we initially believed.  As such, we used
+not running truly in parallel as we initially believed.  As such, we used
 an AMD profiler to report how long each of the kernels took using the static
@@
 benchmark and the expected runtime (that is, the sum of the CPU kernel runtime
-and the GPU kernel runtime).  If the kernels were truely running in parallel,
+and the GPU kernel runtime).  If the kernels were truly running in parallel,
 we would expect to see a negative discrepancy between the actual runtime and
@@
 for GPU kernels.  Our work differs from this, because the kernel execution
-was still being performed soley on the GPU, whereas our work investigates
+was still being performed solely on the GPU, whereas our work investigates
 kernel execution on multiple devices.
@@
 scheduling scheme.  Verner et al. use this approach to create a high-throughput
-low-lantecy heterogeneous encryption system~\cite{Verner2011}.
+low-latency heterogeneous encryption system~\cite{Verner2011}.
 

### 2011-12-05 paper/paper.tex
Added new image for scheduling, and added experimental testbesd section.

@@
 \section{Load Balancing Benchmarks}
-
 For this paper, three different benchmarks were used to assess the
@@
 use of 2 different load balancing schemes in addition to the performance of a CPU-only or
-a GPU-only solution.
+a GPU-only solution.  These schemes are illustrated in Figure~/ref{fig:scheduling}
 
@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=2.5in]{scheduling}
+\caption{Data Partitioning Schemes for Load Balancing.}
+\label{fig:scheduling}
+\end{figure}
+
 \section{Results and Analysis}
+We tested the different benchmark applications using different hardware environments.
+For the discrete GPU tests, we tested our application on an Intel Core Duo CPU with a
+AMD HD5450 GPU attached via PCIe.  For the Fused CPU+GPU environment we used an AMD
+Fusion A-3850.  Linux was used for the discrete environment, while Windows was used
+on the Fused environment (in order to take advantage of newer drivers that allowed
+for better hardware features on the new Fusion machine).
+
 The three implemented benchmarks were run on a system with a discrete GPU, as well as

### 2011-12-05 paper/paper.tex
Added OpenCL section to the paper.

@@
 and data size.
+
+\section{The OpenCL Framework}
+For this project, we will use the OpenCL Framework for our kernel lauching.  OpenCL
+is an application framework for heterogeneous parallel computing.  It includes an
+API for creating and transferring memory to accleration devices, as well as dynamic
+compilation of application kernels that are run on the acceleators.Using OpenCL 
+allows us to write the host code only once, while allowing us to run the core 
+application kernel on the CPU, GPU, or both.
+
+In OpenCL each thread is logically mapped to a work-item.  These work-items are grouped into
+workgroups.  These workgroups map to a SM on a discrete GPU.  For instance, these workgroups
+share access to a SMs local memory and can perform a barrier with all the other work-items
+in the workgroup.  In the case of a CPU acclerator, the workgroups still exist, but optimizations
+such as manual caching via local memory will not be effective.
+
+Devices are grouped into contexts, which contains memory buffers and so on.  A queuing system
+is used by OpenCL to facilitate kernel execution.  Kernels are submitted to the queue along with
+the size of each workgroup, as well as number of workgroups to execute.  This queue is also 
+responsible for reading and writing memory buffers.  The queue is in-order, meaning that 
+the queue will wait until previously enqueued items have finished before starting work 
+on more recently queued items.
+
+For our purposes, we decided to use a seperate context for each device, as well as a seperate
+command queue for each context.  We decided to use a seperate context for each device, as it would
+allow for finer-grained control of data transfers and residency. Using a seperate command queue
+for each device will facilitate multiple devices to be used simultaneously, as each command queue
+operates on its own seperate thread.
 

### 2011-12-05 paper/paper.tex
Typo

--- 2011-12-05 paper/paper.tex
@@
 of any of the benchmarks tested.  There were a few scenarios in which
-a specific static execution ratio achieved meager speedup (<5\%) over
+a specific static execution ratio achieved meager speedup ($<$5\%) over
 just the GPU, but these situations were scarce, and did not achieve the

### 2011-12-05 paper/paper.tex
Increased information in the load-balancing section.

--- 2011-12-05 paper/paper.tex
@@
 the CPU and the GPU at a specified ratio.  This ratio was user-defined allowing for best 
-performance after runtime benchmarking.  The main cause of overhead for this method of
-load-balancing is that of a varying workload.  If the workload for the kernel is imbalanced
+performance after runtime benchmarking (i.e. the precentage of work done on the CPU and GPU).
+The cases of CPU-only and GPU-only scheduling are just special cases of this scheduler, of
+the ratios 0 and 1 respectively.
+
+The static load-balancing scheme benefits from the low overhead of only have to launch the
+kernels on each device only once, as opposed to more complex load-balancing schemes.  This
+increases the amount of work to be done at one time, as workgroups that retire are immediately
+replaced by newer workgroups from the larger queue. The main cause of overhead for this method
+of load-balancing is that of a varying workload.  If the workload for the kernel is imbalanced
 between workgroups, it is impossible to statically determine the proper amount of work
@@
 a dynamic load-balancing scheme.  In this scheme the workgroups were grouped into larger
-chunks.  Each device is initially allocated one chunk of work to work on.  As each chunk
-is retired by the device a new chunk is allocated.  This load-balancing scheme will still
-not guarantee optimal performance.  Within a work chunk, a single long-running workgroup
-can hold up the entire device, not allowing other kernels to execute on open SMs.
+chunks.  Each device is initially allocated one chunk of work to work on, and as each chunk
+is retired by the device a new chunk is allocated.  Each device is given a seperate hardware
+thread, and the thread will wait until the current chunk is finished, then it will synchronize
+through a shared variable to acquire the next chunk of work for the device.  This process
+will continue until the work queue is out of work.
+
+The dynamic load-balancing scheme will still not guarantee optimal performance.
+Within a work chunk, a single long-running workgroup can hold up the entire device,
+not allowing other kernels to execute on open SMs.  This creates and issue where the
+size of the work chunks should be considered.  One one hand, larger work chunks can
+provide for better performance for each device, as it will allow more workgroups to
+hide the cost of a long-running workgroup.  However, large chunk sizes can reduce
+the granularity of load-balancing between the devices.
 

### 2011-12-05 paper/paper.tex
Added some conclusions.

--- 2011-12-05 paper/paper.tex
@@
 \section{Conclusion}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut
-enim ad minim veniam, quis nostrud exercitation ullamco laboris
-nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
-in reprehenderit in voluptate velit esse cillum dolore eu fugiat
-nulla pariatur. Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+It is clear that our original hypotheses were not supported by our results.  That
+is, we were not able to see consistent, significant speedup from any of our load
+balancing schemes.  In many scenarios, the GPU simply outperformed the CPU, with
+the static load balancing schemes falling somewhere between the two, and the dynamic
+load balancer being far worse than anything else.  There were only two notable
+exceptions to this trend.  The first was the reduction benchmark on the discrete GPU
+platform, where the CPU outperformed the GPU due to the low stream processor utilization
+inherent in a GPU reduction implementation.  However, the static load balancing schemes
+were still outperformed by the CPU in this case, so no overall speedup was gained. The
+second exception was the vector add+ benchmark on the discrete GPU platform, where the
+dynamic load balancer actually outperformed the tested static load balancing schemes.
+However, this was due to the vast disparity in the GPU and CPU runtimes for this
+particular benchmark; giving any significant amount of data to the CPU would hurt
+performance.  With the dynamic load balancer, the CPU will only compute a few blocks
+of data, rather than 25\% or more with the static load balancer.
+
+When analyzing more finely-grained static execution ratios, we found that, in the case
+where the GPU outperforms the CPU, there is always a downward trend in overall runtime
+as more data is executed on the GPU.  There were a couple sparse instances where a static
+execution ratio performed better than using just the GPU, but we chalked these up to
+typical variability in runtimes due to their infrequency and minute performance improvements.
+
+When analyzing the results, we considered the possibility that the CPU and GPU kernels were
+actually running in serial, which would explain the performance trends we were seeing in our
+static and dynamic load balancing schemes.  An AMD profiler was used to find out the times each
+kernel took to compute, and the sum of these times was compared to the actual runtime of the
+benchmark.  We found that the actual runtime of the benchmark was typically slightly more than
+the sum of the runtimes of each kernel, which seemed to indicate that the kernels were indeed
+running serially, with some overhead for data transfer times.
+
+Based on the CPU-only and GPU-only runtimes of each benchmark, we were able to calculate an
+optimal static data partitioning, such that the computations on the CPU and GPU took equal
+time to complete.  Based on the disparity of runtimes between the CPU-only and GPU-only
+results, the optimal runtimes ranged from a 2\% to nearly a 45\% speedup over just the GPU
+alone.  Unfortunately,
 

### 2011-12-05 paper/paper.tex
ARRRGGGHHH
Added AMDFusion citation.
Expanded results and conclusion.

@@
 
+With regard to the discrete GPU platform and the fused CPU+GPU platform, it
+seems that the fused platform has some advantages.  For instance, consider 
+Figures \ref{fig:reduction_discrete} and \ref{fig:reduction_fused}.  Although
+we will not compare overall runtimes of the two benchmarks, we can see that
+the GPU of the fused platform performed better than the GPU of the discrete
+platform relative to their CPUs.  That is, the CPU outperformed the GPU on
+the reduction benchmark on the discrete GPU architecture, but the converse
+was true of the fused platform.  This may be due to the fact that the fused
+platform shares memory between the CPU and GPU, which lowers data transfer
+times between the two devices.  On the discrete GPU system, the GPU typically
+experienced significant data transfer times for the GPU, and the effect of
+this would be minimized on the fused platform.
+
 \section{Related Work}
@@
 results, the optimal runtimes ranged from a 2\% to nearly a 45\% speedup over just the GPU
-alone.  Unfortunately,
+alone.  Unfortunately, none of these theoretical results were supported in practice.  It seems
+to be a limitation of OpenCL that it will not run kernels on different devices in parallel,
+which effectively renders our load balancing schemes pointless.
+
+Additionally, it seems that a fused CPU+GPU architecture may provide some performance benefits
+over just a discrete GPU architecture.  Our results shows that the GPU outperformed the CPU on
+the fused platform's reduction benchmark, but this was not the case for the discrete GPU platform.
+However, as the CPUs and GPUs differed significantly between the two testing environments, we
+cannot say definitively that one architecture will always be better than the other.
 
 \begin{thebibliography}{1}
+
+\bibitem{AMDFusion}
+http://fusion.amd.com
 
@@
 
-\setboolean{@twocolumn}{false}
 \newpage

### 2011-12-05 paper/paper.tex
Reworked Load Balancing Benchmarks.

--- 2011-12-05 paper/paper.tex
@@
 For this paper, three different benchmarks were used to assess the
-effectiveness of the load balancer.  The first is a reduction benchmark,
-in which the sum of variably-sized array is computed.  The reduction benchmark
-is unique in that different code is being run on the CPU and the GPU.  The
-reason for this is that the GPU cannot efficiently compute a reduction the way
-a CPU typically would -- due to the relative inefficiency of backward branches
-on the GPU.  A reduction on the GPU is a multi-stage process, in which the
-output of one step is used as the input of the next.  Each steam processor
-computes the reduction of two elements during each iteration, reducing the amount
-of work during each step by a factor of two.  As a result, each step utilizes half
-the stream processors of the previous step.  This continues until the reduction
-operation is only done by a single stream processor.  A natural consequence of
-this algorithm is that not all of the stream processors will be utilized during
-the computation.
+effectiveness of the load balancer.  These benchmarks were chosen to represent
+tasks that would actually be performed on a GPU, and have seen use in already
+existing benchmark suites~\cite{Danalis2010}.  The applications that we have chosen
+for this project are VectorAdd, VectorAdd+, and Reduce.
 
-The next benchmark is a simple vector addition operation that computes the sum
-of two input vectors, and stores the result into an output vector.  The third
-benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
-the Vector Add benchmark, but it computes 1000 additions before storing the result
+The VectorAdd application is a simple vector addition operation that computes the sum
+of two input vectors, and stores the result into an output vector.  This represents the
+vector operation $\vec{C} = \vec{A} + \vec{B}$.  This kernel is naive to parallelize, as
+it exhibits embarrassingly parallel structure.  Each thread being run maps to a single 
+index in the array and computes the output value for that index.  This algorithm is also
+nicely designed for load-balancing because of the lack of data dependencies, reducing 
+the amount of communication to the GPU.
+
+The next benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
+the VectorAdd benchmark, but it computes 1000 additions before storing the result
 in the output vector.  The motivation behind this benchmark is to reduce memory
-access overhead with more computation.  All numbers reported in the results section
+access overhead with more computation.  This benchmark was compiled without optimizations
+so that the additions would not be optmized into a single multiplication.  The data
+dependencies are exactly the same as for the VectorAdd application, so the results for
+data transfer speeds should be equal for both applications, and the difference between
+the two should only be the kernel execution times.
+
+The Reduce application is an application that performs a reduction operation on a
+large array of data.  In this case, the sum of a variably-sized array is computed.
+The Reduce benchmark is unique in that different code is being run on the CPU and 
+the GPU.  The reason for this is that the GPU cannot efficiently compute a reduction
+the way a CPU typically would -- due to the relatively slower cores and much greater
+memory latency.  A optimized reduction on the GPU is a multi-stage process, in which the
+output of one step is used as the input of the next.  Each SM computes the reduction
+of a smaller set of elements from the larger array during each iteration.  This
+requires the use of numerous local atomic operations, as well as local memory to 
+facilitate optimized memory access between threads on an SM.
+This process continues until the reduction operation is only done by a single SM.
+A natural consequence of this algorithm is that not all of the SMs will be utilized during
+the computation.  In addition, inside an iteration all of the streaming processors
+will only be used for the first stage of reduction, greatly reducing utilization on the
+device.
+
+All numbers reported in the results section
 are given as a mean of ten runs for that particular benchmark, load-balancing scheme,
@@
 
+\bibitem{AMDFusion}
+http://fusion.amd.com
+
 \bibitem{Augonnet2009}

### 2011-12-05 paper/paper.tex

@@
 results, the optimal runtimes ranged from a 2\% to nearly a 45\% speedup over just the GPU
-alone.  Unfortunately,
+alone.  Unfortunately, none of these theoretical results were supported in practice.  It seems
+to be a limitation of OpenCL that it will not run kernels on different devices in parallel,
+which effectively renders our load balancing schemes pointless.
+
+Additionally, it seems that a fused CPU+GPU architecture may provide some performance benefits
+over just a discrete GPU architecture.  Our results shows that the GPU outperformed the CPU on
+the fused platform's reduction benchmark, but this was not the case for the discrete GPU platform.
+However, as the CPUs and GPUs differed significantly between the two testing environments, we
+cannot say definitively that one architecture will always be better than the other.
 

### 2011-12-05 paper/paper.tex
Improved the results section of the paper.

@@
 Fusion A-3850.  Linux was used for the discrete environment, while Windows was used
-on the Fused environment (in order to take advantage of newer drivers that allowed
-for better hardware features on the new Fusion machine).
+on the Fused environment in order to take advantage of newer drivers that allowed
+for better hardware features on the new Fusion machine.
 
@@
 due to the GPU not utilizing all of its stream processors during a reduction operation, as
-discussed previously.  Unfortunately, none of the static load balancing schemes were able
-to outperform just the CPU; all the static times were somewhere in between the CPU-only and
-GPU-only times.  Additionally, the dynamic load balancing scheme performed significantly
-worse than all the other schemes, and more than 100\% worse than just the CPU.
-100\% worse than just the CPU.
+discussed previously.  It is also the case that the large amount of synchronization that must
+occur in each iteration of the GPU reduction, and the number of kernel launches required for
+each iteration that is not present in the CPU reduction. In addition, because the CPU had better
+performance on the kernel alone, it saw an even greater improvement when the data transfer costs
+were considered, with the CPU-only case requiring no data movement at all. Unfortunately, none of
+the static load balancing schemes were able to outperform the CPU-only scheme; all the static times were
+somewhere in between the CPU-only and GPU-only times.  Additionally, the dynamic load balancing
+scheme performed significantly worse than all the other schemes, and more than 100\%
+worse than just the CPU.  The cause for this degredation in performance was likely due to the
+overhead of scheduling threads and communciation between those threads, as well as moving data to
+the GPU in small chunks when the GPU showed very poor performance compared to the CPU.
 
@@
 
-Now, consider Figure \ref{fig:vector_discrete}, where the vector add benchmark was run on
+Now, consider Figure \ref{fig:vector_discrete}, where the VectorAdd benchmark was run on
 the discrete GPU platform.  In this case, we see that the GPU has nearly a 4x speedup over
-just the CPU.  However, similar to the reduction benchmark, the static load balancing schemes
-never outperform just the GPU alone, with all of them lying somewhere between the CPU-only
+just the CPU.  The data parallelism and lack of any synchronization requirements makes this
+speedup over the CPU possible.  However, similar to the reduction benchmark, the static load balancing schemes
+never outperform just the GPU-only scheme, with all of them lying somewhere between the CPU-only
 and GPU-only runs.  Also similar to the reduction benchmark, the dynamic load balancing scheme
@@
 
-The results for the vector add+ benchmark on the discrete GPU platform are shown in Figure
+The results for the VectorAdd+ benchmark on the discrete GPU platform are shown in Figure
 \ref{fig:vector_plus_discrete}.  There are a few notable trends in this figure that are not
 present in the previous two.  The first is is how drastically the GPU outperforms the CPU --
-with over a 100x speedup over just the CPU with 16~MB of data.  The second is that the
+with over a 100x speedup over just the CPU with 16~MB of data.  Because of the very large
+amount of computation that needs to be performed, compared to the low amount of data accesses
+required in the kernel (3 data accesses for 1000 additions), the GPU is able to outperform
+the CPU because of its very large throughput for computationally intensive kernels.  The second 
+thing to notice for this benchmark specificially is that the
 dynamic load balancing drastically outperforms the static load balancing schemes.  This is
@@
 balancing scheme, only small chunks of data are given to the CPU at a time, rather than a
-static percentage.
+static percentage, allowing for the GPU to take advantage of much more of its speedup over
+the CPU. Increasing the chunk size for the GPU (taking two chunks instead of one at a time)
+would likely yield in a greater speedup for the dynamic load-balancing case.
 
@@
 data ratios run on the GPU -- 25\%, 50\%, and 75\%.  The general trend was that
-as more of the workload was placed on the GPU, the runtime improved.  As a
+as more of the workload was placed on the GPU, the runtime improved, with
+an exception in the Reduce benchmark due to the superior CPU performance.  As a
 result, we run the vector add benchmark again with a high granularity of data
@@
 325.27 seconds, compared to 333.31 seconds with only the GPU -- a ~2.4\%
-speedup.  However, these few instances of speedup may simply be due to variability.
+speedup.  However, these few instances of speedup may simply be due to variability in
+the system.
 
@@
 \ref{fig:vector_plus_fused}. The figures demonstrate a variety of different
-load balancing options for the Reduction, Vector Add, and Vector Add+
-benchmarks, across data sizes ranging between 1~MB and 16~MB.  Each figure
-depicts the total execution time for the CPU-only and GPU-only computations,
+load balancing options for the Reduction, VectorAdd, and VectorAdd+
+benchmarks, across data sizes ranging between 1~MB and 16~MB on a Fused CPU+GPU platform.
+Each figure depicts the total execution time for the CPU-only and GPU-only computations,
 as well as static load balancing with 25\%, 50\%, and 75\% of the data on the
@@
 faster than the CPU for the given benchmarks, with the static partitioning
-schemes lying somewhere between the two.
+schemes lying somewhere between the two.  As you can see in the figures, while the
+overall performance numbers have changed between the discrete and fused architectures
+the major trends from the previous graphs still hold true.
 
@@
 (0.5\% and 4.4\% respective speedups for 9~MB, 1.5\% and 3.6\% respective speedups
-for 10~MB). What is also interested to note is that on the fused CPU+GPU architecture,
+for 10~MB). What is also interesting to note is that on the fused CPU+GPU architecture,
 the GPU outperformed the CPU on the reduction benchmark.  However, the opposite was
-true of the reduction benchmark on the discrete GPU platform.
+true of the reduction benchmark on the discrete GPU platform.  The difference in this
+performance was likely due to the increased transfer speeds between the CPU and GPU
+on the fused platform.  It is also the case that the SIMD engine on the Fusion APU
+is slightly more powerful than the discrete GPU that we have used for testing, allowing
+kernel execution times on the APU to be lower than that of the discrete GPU.
 
@@
 not running truly in parallel as we initially believed.  As such, we used
-an AMD profiler to report how long each of the kernels took using the static
+the OpenCL profiling capabilities to report how long each of the kernels took using the static
 load balancer with 95\% of the data on the CPU.  Figure \ref{fig:sequential_discrepancy}
 shows the discrepancy between the actual total runtime of the vector add
-benchmark and the expected runtime (that is, the sum of the CPU kernel runtime
+benchmark and the expected runtime of sequential calls to the kernels 
+(that is, the sum of the CPU kernel runtime
 and the GPU kernel runtime).  If the kernels were truly running in parallel,
 we would expect to see a negative discrepancy between the actual runtime and
-expected runtime.
+expected runtime.  It was impossible for us to simply time the start and end
+times of the different kernels, because while that information was available
+for each kernel launch, it was kept as device-speciifc times. We could not
+reconcile the two different timing schemes, because of the different clock
+frequencies and the lack of a mapping from time from clock start for either
+device.
 
@@
 seems that the fused platform has some advantages.  For instance, consider 
-Figures \ref{fig:reduction_discrete} and \ref{fig:reduction_fused}.  Although
+Figures \ref{fig:reduce_discrete} and \ref{fig:reduce_fused}.  Although
 we will not compare overall runtimes of the two benchmarks, we can see that

### 2011-12-05 paper/paper.tex
Added the appendix about sharing work throughout the project.

@@
 \appendix[Team Member Contributions]
-Kenneth Lee did A, B, C.  Kyle Morgan did X, Y, Z.
+{\bf Kenneth Lee} was responsible for load-balancing schemes in
+both design and implementation.  He was also in charge of system
+integration and testing of the final applications on the appropriate
+hardware devices.  Kenneth was also in charge of deep analysis and
+explanation of result data.
+
+{\bf Kyle Morgan} was in charge of kernel design and implementation.
+In addition, he was also in charge of result compilation and 
+preliminary analysis.  Kyle was also responsible for background readings
+in both GPGPU computation in general, as well as specifics in terms of
+both GPU architecture and the OpenCL framework.
+
 \end{document}

### 2011-12-05 paper/paper.tex
Various typo changes.

--- 2011-12-05 paper/paper.tex
@@
 For this paper, three different benchmarks were used to assess the
-effectiveness of the load balancer.  These benchmarks were chosen to represent
-tasks that would actually be performed on a GPU, and have seen use in already
-existing benchmark suites~\cite{Danalis2010}.  The applications that we have chosen
-for this project are VectorAdd, VectorAdd+, and Reduce.
+effectiveness of the load balancer.  The first is a reduction benchmark,
+in which the sum of variably-sized array is computed.  The reduction benchmark
+is unique in that different code is being run on the CPU and the GPU.  The
+reason for this is that the GPU cannot efficiently compute a reduction the way
+a CPU typically would -- due to the relative inefficiency of backward branches
+on the GPU.  A reduction on the GPU is a multi-stage process, in which the
+output of one step is used as the input of the next.  Each steam processor
+computes the reduction of two elements during each iteration, reducing the amount
+of work during each step by a factor of two.  As a result, each step utilizes half
+the stream processors of the previous step.  This continues until the reduction
+operation is only done by a single stream processor.  A natural consequence of
+this algorithm is that not all of the stream processors will be utilized during
+the computation.
 
-The VectorAdd application is a simple vector addition operation that computes the sum
-of two input vectors, and stores the result into an output vector.  This represents the
-vector operation $\vec{C} = \vec{A} + \vec{B}$.  This kernel is naive to parallelize, as
-it exhibits embarrassingly parallel structure.  Each thread being run maps to a single 
-index in the array and computes the output value for that index.  This algorithm is also
-nicely designed for load-balancing because of the lack of data dependencies, reducing 
-the amount of communication to the GPU.
-
-The next benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
-the VectorAdd benchmark, but it computes 1000 additions before storing the result
+The next benchmark is a simple vector addition operation that computes the sum
+of two input vectors, and stores the result into an output vector.  The third
+benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
+the Vector Add benchmark, but it computes 1000 additions before storing the result
 in the output vector.  The motivation behind this benchmark is to reduce memory
-access overhead with more computation.  This benchmark was compiled without optimizations
-so that the additions would not be optmized into a single multiplication.  The data
-dependencies are exactly the same as for the VectorAdd application, so the results for
-data transfer speeds should be equal for both applications, and the difference between
-the two should only be the kernel execution times.
-
-The Reduce application is an application that performs a reduction operation on a
-large array of data.  In this case, the sum of a variably-sized array is computed.
-The Reduce benchmark is unique in that different code is being run on the CPU and 
-the GPU.  The reason for this is that the GPU cannot efficiently compute a reduction
-the way a CPU typically would -- due to the relatively slower cores and much greater
-memory latency.  A optimized reduction on the GPU is a multi-stage process, in which the
-output of one step is used as the input of the next.  Each SM computes the reduction
-of a smaller set of elements from the larger array during each iteration.  This
-requires the use of numerous local atomic operations, as well as local memory to 
-facilitate optimized memory access between threads on an SM.
-This process continues until the reduction operation is only done by a single SM.
-A natural consequence of this algorithm is that not all of the SMs will be utilized during
-the computation.  In addition, inside an iteration all of the streaming processors
-will only be used for the first stage of reduction, greatly reducing utilization on the
-device.
-
-All numbers reported in the results section
+access overhead with more computation.  All numbers reported in the results section
 are given as a mean of ten runs for that particular benchmark, load-balancing scheme,
@@
 use of 2 different load balancing schemes in addition to the performance of a CPU-only or
-a GPU-only solution.  These schemes are illustrated in Figure~/ref{fig:scheduling}
+a GPU-only solution.  These schemes are illustrated in Figure~\ref{fig:scheduling}.
 
@@
 worse than all the other schemes, and more than 100\% worse than just the CPU.
-100\% worse than just the CPU.
 
@@
 However, this was due to the vast disparity in the GPU and CPU runtimes for this
-particular benchmark; giving any significant amount of data to the CPU would hurt
+particular benchmark, as giving any significant amount of data to the CPU would hurt
 performance.  With the dynamic load balancer, the CPU will only compute a few blocks
@@
 to be a limitation of OpenCL that it will not run kernels on different devices in parallel,
-which effectively renders our load balancing schemes pointless.
+which effectively renders our current load balancing schemes pointless.
 
 Additionally, it seems that a fused CPU+GPU architecture may provide some performance benefits
-over just a discrete GPU architecture.  Our results shows that the GPU outperformed the CPU on
+over just a discrete GPU architecture.  Our results show that the GPU outperformed the CPU on
 the fused platform's reduction benchmark, but this was not the case for the discrete GPU platform.

### 2011-12-05 paper/paper.tex
Included more figures.

--- 2011-12-05 paper/paper.tex
@@
 hide the cost of a long-running workgroup.  However, large chunk sizes can reduce
-the granularity of load-balancing between the devices.
+the granularity of load-balancing between the devices.  Alternatively, smaller chunk
+sizes alleviates the problem of one device receiving a disproportionate amount of the
+work, but can increase scheduling overhead by increasing contention for the shared
+variable which tracks the remaining workload.
 
@@
 
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=3.0in]{reduce_static_fused}
-%\caption{Reduction Benchmark with Varied Execution Ratio}
-%\label{fig:reduce_static_fused}
-%\end{figure}
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{reduce_static_fused}
+\caption{Reduction Benchmark with Varied Execution Ratio}
+\label{fig:reduce_static_fused}
+\end{figure}
 
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=3.0in]{vector_static_fused}
-%\caption{Vector Add Benchmark with Varied Execution Ratio}
-%\label{fig:vector_static_fused}
-%\end{figure}
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_static_fused}
+\caption{Vector Add Benchmark with Varied Execution Ratio}
+\label{fig:vector_static_fused}
+\end{figure}
 
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=3.0in]{vector_plus_static_fused}
-%\caption{Vector Add+ Benchmark with Varied Execution Ratio}
-%\label{fig:vector_plus_static_fused}
-%\end{figure}
+\begin{figure}[t]
+\centering
+\includegraphics[width=3.0in]{vector_plus_static_fused}
+\caption{Vector Add+ Benchmark with Varied Execution Ratio}
+\label{fig:vector_plus_static_fused}
+\end{figure}
 
 %Finally, consider Figures \ref{fig:reduce_static_fused}, \ref{fig:vector_static_fused},
-%and \ref{fig:vector_plus_static_fused}. In each, the runtim for both the CPU and GPU are
+%and \ref{fig:vector_plus_static_fused}. In each, the runtime for both the CPU and GPU are
 %shown with a 16~MB problem size.  Additionally, a variety of static execution ratios are
@@
 %able to achieve a better time than just the GPU.
+
+Finally, consider Figure \ref{fig:vector_static_fused}, where the overall runtime
+for the static load balancer is compared against both the CPU-only and GPU-only
+runtimes for 0.1\%, 1\%, 10\%, 25\%, 50\%, 75\%, 99\%, and 99.9\% static data
+ratios for the 16~MB vector add benchmark.  Figures \ref{fig:reduce_static_fused}
+and \ref{fig:vector_plus_static_fused} show similar results for the reduction and
+vector add+ benchmarks.  In these benchmarks, the downward trend as more data is
+executed on the GPU becomes more evident.  What's also interesting is that the
+static load balancer actually performs worse than the CPU-only case when only a
+small amount of the data ($\leq$1\%) is executed on the CPU in the reduction and
+vector add benchmarks.
 

### 2011-12-05 paper/paper.tex
Added to APU section.

--- 2011-12-05 paper/paper.tex
@@
 Fusion technology is an example of this type of architecture~\cite{AMDFusion}.
+Because of the limited space on the CPU die, the overall performance
+capabilities of the CPU and GPU are often more limited than on a traditional 
+discrete CPU and GPU architecture.
 
@@
 between CPU and GPU, allowing for an overall improvement in runtime, despite
-reduced compute power on the GPU~\cite{Daga2011}.
+reduced compute power on the GPU~\cite{Daga2011}.  The improvment of the 
+data transfer times greatly improves the applicability of the GPU for use
+in general computing paradigms, specifically that of CPU+GPU co-processing.
+Decreasing the data trasnsfer time reduces the penalty associated with migrating
+computation to the GPU, and therefore can access the computation of the GPU
+more at will then in discrete GPUs which are bound by the slow access speeds
+of the PCI-e bus.
 
@@
 actually running in serial, which would explain the performance trends we were seeing in our
-static and dynamic load balancing schemes.  An AMD profiler was used to find out the times each
+static and dynamic load balancing schemes.  An OpenCL profiler was used to find out the times each
 kernel took to compute, and the sum of these times was compared to the actual runtime of the

### 2011-12-05 paper/paper.tex
Added to GPU and CPU architectures.

--- 2011-12-05 paper/paper.tex
@@
 machines under test.  We will discuss CPU, GPU, and APU architectures in
-terms of major differences that affect the performance of the device.
+terms of major differences that affect the performance of the device.  We also
+briefly discuss the impacts of different architectures on the implementations
+of a CPU+GPU co-processing system.
 
@@
 
+In addition, the CPU typically has a deep pipeline to increase throughput of the system.
+Longer and longer pipelines increase the utilization of every element found on the CPU,
+further increasing latency-critical performance.  Performance of this pipeline is in danger
+of being flushed on braching codes, but advanced branch prediction and speculative execution
+are able to greatly reduce the amount of times where a pipeline flush must occur.  Out-of-order
+execution of instructions in the program also improves performance for single threads on the
+CPU.
+
 The use of the Single Instruction Multiple Data (SIMD) paradigm enables increased
@@
 applications.  However, optimal performance of these types of instructions often
-involves complex compilation techniques and hand-tuning of the assembly.
+involves complex compilation techniques and hand-tuning of the assembly.  Most applications
+are not able to see as substantial a speedup from this process as they would if the application
+were really able to process 4 threads at once.
 
@@
 
+Each processor on the GPU is very dumb and primitive compared to the processors on the CPU.
+There is no speculative execution, deep pipelines, out-of-order exeuction, or branch prediction
+on this device.  Instead, the space saved by not having these features is instead used to
+cram more and more compute cores onto the device.  This strategy of having a large number of
+very simple cores goes back to the orginial architectual decisions for the GPU.  Graphics applications
+are relatively straight-forward and require a lot of math and very little complex coding.
+Therefore, for graphics, there is little performance lost for not having these features, whereas
+for GPGPU applications these features can greatly diminish overall kernel performance.
+
 Groups of processing units, Streaming Multiprocessors (SMs), share a single instruction
@@
 features divergent branching in a warp (caused by if-else clauses) greatly degrade application
-performance.
+performance.  Unlike the CPU, the GPU is unable to avoid warp-level divergence of
+the different threads, and therefore must always take the penalty of braching.
 

### 2011-12-05 paper/paper.tex
Added to Conclusions.

@@
 
+If time had permitted it would have been beneficial to analyze the effect of different
+data chunk sizes on the overall performance of the dynamic scheduler, including having
+different data chunk sizes for each device, based on computational power.  We could have
+also included a ratio-based dynamic scheduler which would have pulled a percentage of the
+remaining data from the work-queue instead of a set static amount of data.  This would have
+alleviated the problem of not having enough work to do at the outset of the problem, while
+still retaining fine-grained distribution of data near the end of the queue.  The lack of
+this sort of analysis limits our overall ability to throughly analyze the performance
+of the dynamic scheduler.
+
 When analyzing more finely-grained static execution ratios, we found that, in the case

### 2011-12-05 paper/paper.tex
Spelling fixes.

@@
 further increasing latency-critical performance.  Performance of this pipeline is in danger
-of being flushed on braching codes, but advanced branch prediction and speculative execution
+of being flushed on branching codes, but advanced branch prediction and speculative execution
 are able to greatly reduce the amount of times where a pipeline flush must occur.  Out-of-order
@@
 Each processor on the GPU is very dumb and primitive compared to the processors on the CPU.
-There is no speculative execution, deep pipelines, out-of-order exeuction, or branch prediction
+There is no speculative execution, deep pipelines, out-of-order execution, or branch prediction
 on this device.  Instead, the space saved by not having these features is instead used to
 cram more and more compute cores onto the device.  This strategy of having a large number of
-very simple cores goes back to the orginial architectual decisions for the GPU.  Graphics applications
+very simple cores goes back to the original architectural decisions for the GPU.  Graphics applications
 are relatively straight-forward and require a lot of math and very little complex coding.
@@
 performance.  Unlike the CPU, the GPU is unable to avoid warp-level divergence of
-the different threads, and therefore must always take the penalty of braching.
+the different threads, and therefore must always take the penalty of branching.
 
@@
 between CPU and GPU, allowing for an overall improvement in runtime, despite
-reduced compute power on the GPU~\cite{Daga2011}.  The improvment of the 
+reduced compute power on the GPU~\cite{Daga2011}.  The improvement of the 
 data transfer times greatly improves the applicability of the GPU for use
 in general computing paradigms, specifically that of CPU+GPU co-processing.
-Decreasing the data trasnsfer time reduces the penalty associated with migrating
+Decreasing the data transfer time reduces the penalty associated with migrating
 computation to the GPU, and therefore can access the computation of the GPU
@@
 access overhead with more computation.  This benchmark was compiled without optimizations
-so that the additions would not be optmized into a single multiplication.  The data
+so that the additions would not be optimized into a single multiplication.  The data
 dependencies are exactly the same as for the VectorAdd application, so the results for
@@
 \section{The OpenCL Framework}
-For this project, we will use the OpenCL Framework for our kernel lauching.  OpenCL
+For this project, we will use the OpenCL Framework for our kernel launching.  OpenCL
 is an application framework for heterogeneous parallel computing.  It includes an
-API for creating and transferring memory to accleration devices, as well as dynamic
-compilation of application kernels that are run on the acceleators.Using OpenCL 
+API for creating and transferring memory to acceleration devices, as well as dynamic
+compilation of application kernels that are run on the accelerators.Using OpenCL 
 allows us to write the host code only once, while allowing us to run the core 
@@
 share access to a SMs local memory and can perform a barrier with all the other work-items
-in the workgroup.  In the case of a CPU acclerator, the workgroups still exist, but optimizations
+in the workgroup.  In the case of a CPU accelerator, the workgroups still exist, but optimizations
 such as manual caching via local memory will not be effective.
@@
 
-For our purposes, we decided to use a seperate context for each device, as well as a seperate
-command queue for each context.  We decided to use a seperate context for each device, as it would
-allow for finer-grained control of data transfers and residency. Using a seperate command queue
+For our purposes, we decided to use a separate context for each device, as well as a separate
+command queue for each context.  We decided to use a separate context for each device, as it would
+allow for finer-grained control of data transfers and residency. Using a separate command queue
 for each device will facilitate multiple devices to be used simultaneously, as each command queue
-operates on its own seperate thread.
+operates on its own separate thread.
 
@@
 the CPU and the GPU at a specified ratio.  This ratio was user-defined allowing for best 
-performance after runtime benchmarking (i.e. the precentage of work done on the CPU and GPU).
+performance after runtime benchmarking (i.e. the percentage of work done on the CPU and GPU).
 The cases of CPU-only and GPU-only scheduling are just special cases of this scheduler, of
@@
 chunks.  Each device is initially allocated one chunk of work to work on, and as each chunk
-is retired by the device a new chunk is allocated.  Each device is given a seperate hardware
+is retired by the device a new chunk is allocated.  Each device is given a separate hardware
 thread, and the thread will wait until the current chunk is finished, then it will synchronize
@@
 scheme performed significantly worse than all the other schemes, and more than 100\%
-worse than just the CPU.  The cause for this degredation in performance was likely due to the
-overhead of scheduling threads and communciation between those threads, as well as moving data to
+worse than just the CPU.  The cause for this degradation in performance was likely due to the
+overhead of scheduling threads and communication between those threads, as well as moving data to
 the GPU in small chunks when the GPU showed very poor performance compared to the CPU.
@@
 the CPU because of its very large throughput for computationally intensive kernels.  The second 
-thing to notice for this benchmark specificially is that the
+thing to notice for this benchmark specifically is that the
 dynamic load balancing drastically outperforms the static load balancing schemes.  This is
@@
 times of the different kernels, because while that information was available
-for each kernel launch, it was kept as device-speciifc times. We could not
+for each kernel launch, it was kept as device-specific times. We could not
 reconcile the two different timing schemes, because of the different clock

### 2011-12-05 paper/paper.tex
Polish.

@@
 major problems facing computer scientists today.  General-purpose computing
-on graphics processing units (GPGPU) is a technique that is becoming
+on Graphics Processing Units (GPGPU) is a technique that is becoming
 increasingly popular for handling computations typically performed by the
 CPU.  However, many GPGPU algorithms place the entire workload on the Graphics
-processing unit (GPU) leaving the CPU idle for the duration of the computation.
+Processing Unit (GPU) leaving the CPU idle for the duration of the computation.
 This paper investigates the effect various load-balancing schemes across the CPU
 and GPU have on overall application performance.  Additionally, the discrete and
-fused GPU architectures, and their impact on performance, will be investigated.
+fused GPU architectures, and their impact on performance, has been investigated.
 \end{abstract}
@@
 GPU gives it a place as a massively parallel hardware accelerator.  Programming frameworks
-such as OpenCL allow programmers to utilize the stream processors of the GPU for non-graphics
-data.  However, because of the difficulty in programming for this hardware, most of the
-effort is spent on trying to achieve performance from the GPU, while the CPU is left
-unused.  By also making use of CPU to process a chunk of the data, we can achieve higher
+such as OpenCL and CUDA allow programmers to utilize the stream processors of the GPU for non-graphics
+applications.  However, because of the difficulty in programming for this hardware, most of the
+attention is spent on trying to achieve optimal performance on the GPU to leverage the unique hardware
+capabilites, while the CPU is left
+unused.  By also making use of CPU to process a portion of the overall workload, we are able to achieve higher
 utilization of resources and therefore increase overall program performance.
@@
 transfer cost, and $t_o$ is any overhead incurrent from synchronization costs of the load
-balancing scheme or otherwise.  The above will be faster than just the GPU iff
+balancing scheme or otherwise.  Note that there is an obvious minimum for this function the
+point where the GPU and CPU workloads are equal.  The above will be faster than just the GPU iff
 $t_g - \max{\{(n-m)t_g, mt_c}\} < t_d + t_o$.
@@
 of being flushed on branching codes, but advanced branch prediction and speculative execution
-are able to greatly reduce the amount of times where a pipeline flush must occur.  Out-of-order
+are able to greatly reduce the amount of times when a pipeline flush must occur.  Out-of-order
 execution of instructions in the program also improves performance for single threads on the
@@
 architecture of the device to increase performance of a single thread, the GPU
-architecture is specifically designed for large throughput performance.  Large
+architecture is specifically designed for high-throughput performance.  Large
 numbers of threads can be easily created an run on a GPU.  Large register files
@@
 performance the latency to memory is instead hidden through aggressive context
-switching.  Newer GPU architectures may contain privative caching schemes.
+switching.  Newer GPU architectures may contain privative caching schemes, but the majority
+of GPUs still on the commodity graphics market do not have this feature.
 Each SM on the GPU also contains a local memory memory space which can accessed 
@@
 API for creating and transferring memory to acceleration devices, as well as dynamic
-compilation of application kernels that are run on the accelerators.Using OpenCL 
+compilation of application kernels that are run on the accelerators. Using OpenCL 
 allows us to write the host code only once, while allowing us to run the core 
@@
 
-In OpenCL each thread is logically mapped to a work-item.  These work-items are grouped into
+In OpenCL each thread is logically mapped to a so-called ``work-item''.  These work-items are grouped into
 workgroups.  These workgroups map to a SM on a discrete GPU.  For instance, these workgroups
@@
 
-Devices are grouped into contexts, which contains memory buffers and so on.  A queuing system
+Devices are grouped into contexts, which contains memory buffers and device specific information.  A queuing system
 is used by OpenCL to facilitate kernel execution.  Kernels are submitted to the queue along with
@@
 For our purposes, we decided to use a separate context for each device, as well as a separate
-command queue for each context.  We decided to use a separate context for each device, as it would
-allow for finer-grained control of data transfers and residency. Using a separate command queue
+command queue for each context.  Using a separate command queue
 for each device will facilitate multiple devices to be used simultaneously, as each command queue
-operates on its own separate thread.
+operates on its own separate thread. We also decided to use a separate context for each device, as it would
+allow for finer-grained control of data transfers and residency.
 
@@
 
+\subsection{Discrete GPU Performance}
 First, the reduction benchmark was run on the discrete GPU platform with a variety of
@@
 
+\subsection{Fused APU Performance}
 Next, consider Figures \ref{fig:reduce_fused}, \ref{fig:vector_fused}, and
@@
 
-Finally, consider Figure \ref{fig:vector_static_fused}, where the overall runtime
+\subsection{Load Balancing Performance}
+Consider Figure \ref{fig:vector_static_fused}, where the overall runtime
 for the static load balancer is compared against both the CPU-only and GPU-only
@@
 
+Additionally, the use of only applications which have a static workload per thread limit the
+effectiveness of the dynamic scheduler.  The ability to a priori know the thread workload makes
+the static load-balancing scheme more effective than the dynamic scheme.
+Using a kernel with a varied workload per thread
+would have made it possible for the dynamic scheduler to see improved performance when
+compared to the static mapping of data.
+
 When analyzing more finely-grained static execution ratios, we found that, in the case
